---
title: "Integrate RNA"
author: "Michael Gildea"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: true
    df_print: kable
    code_folding: hide
params:
  merged_rds_path: "merged_QC.rds"
  integrated_rds_path: "integrated.rds"
  integrated_sct_rds_path: "integrated_SCT.rds"
  output_dir: "."
  integration_level: "Batch"
  lisi_rds_path: "lisi_integrated_RNA.rds"
  lisi_sct_rds_path: "lisi_integrated_SCT.rds"
  markers_csv_path: "FindMarkersBulk_RNA_integrated/Top_markers.csv"
  markers_sct_csv_path: "FindMarkersBulk_SCT_integrated/Top_markers.csv"
---

<style type="text/css">
  body{
  font-size: 12pt;
}
  h1{
  font-size: 40pt;
}
  h2{
  font-size: 30pt;
}
  h3,h4,h5,h6{
  font-size: 20pt;
}
</style>

Report metadata: print knit timestamp and R version for reproducibility.
```{r report-metadata}
time <- Sys.time()
print(paste('This notebook was knit on:', time))
print(version)
```

Environment context: print working directory used during render.
```{r working-directory}
cat(paste('Bigpurple Location: ',getwd()))
```

Environment package snapshot: attempt to print conda package list for debugging.
```{r conda-package-list}
try(system('conda list'), silent = TRUE)
```

Load required R packages for integration, plotting, and diagnostics.
```{r load-libraries, message=F, warning=F}
library(Seurat)
library(ggplot2)
library(pheatmap)
library(lisi)
library(CVRCFunc)
```

Parameter validation: check integration level and resolve output paths.
```{r validate-params-and-paths, warning=F, message=F}
if (!params$integration_level %in% c('Batch', 'Sample')) {
  stop("params$integration_level must be 'Batch' or 'Sample'.")
}

integration_group_column <- if (params$integration_level == 'Batch') 'batch' else 'sample_name'

resolve_output_path <- function(path_val) {
  if (grepl('^/', path_val)) return(path_val)
  file.path(params$output_dir, path_val)
}

lisi_rds_path <- resolve_output_path(params$lisi_rds_path)
markers_csv_path <- resolve_output_path(params$markers_csv_path)
lisi_sct_rds_path <- resolve_output_path(params$lisi_sct_rds_path)
markers_sct_csv_path <- resolve_output_path(params$markers_sct_csv_path)
```

Track run progress by logging each code chunk as it starts, with timestamps.
```{r chunk-progress-logging, include=FALSE, warning=F, message=F }
safe_level <- gsub('[^A-Za-z0-9._-]+', '_', params$integration_level)
chunk_log_dir <- file.path(params$output_dir, 'chunk_logs')
dir.create(chunk_log_dir, showWarnings = FALSE, recursive = TRUE)
progress_log <- file.path(chunk_log_dir, paste0('chunk_progress_integrate_RNA_', safe_level, '.log'))
cat(sprintf('\n=== integrate_RNA run started (%s): %s ===\n',
            params$integration_level,
            format(Sys.time(), '%Y-%m-%d %H:%M:%S')),
    file = progress_log,
    append = TRUE)

knitr::opts_hooks$set(label = function(options) {
  chunk_label <- if (is.null(options$label) || !nzchar(options$label)) {
    'unnamed-chunk'
  } else {
    options$label
  }

  line <- sprintf('[%s] START chunk: %s\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), chunk_label)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  options
})

knitr::knit_hooks$set(document = function(x) {
  line <- sprintf('[%s] END render: integrate_RNA (%s)\n',
                  format(Sys.time(), '%Y-%m-%d %H:%M:%S'),
                  params$integration_level)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  x
})
```

Input loading: read merged object and verify required integration metadata exists.
```{r load-merged-object, warning=F, message=F }
if (!file.exists(params$merged_rds_path)) {
  stop('Merged RDS not found: ', params$merged_rds_path)
}
seurat <- readRDS(params$merged_rds_path)

if (!integration_group_column %in% colnames(seurat@meta.data)) {
  stop('Missing integration metadata column: ', integration_group_column)
}
```

Integration workflow: split by selected level, integrate RNA, cluster, and save output RDS.
```{r integrate-log-norm, warning=F, message=F }
print(paste('Integrating RNA at level:', params$integration_level))
options(future.globals.maxSize = 8000 * 1024^8)

seurat[["RNA"]] <- split(
  seurat[["RNA"]],
  f = seurat@meta.data[[integration_group_column]]
)
seurat <- NormalizeData(seurat, assay = 'RNA')
seurat <- FindVariableFeatures(seurat, assay = 'RNA')
seurat <- ScaleData(seurat, assay = 'RNA')
seurat <- RunPCA(seurat, assay = 'RNA', reduction.name = 'PCA_RNA_split')
seurat <- IntegrateLayers(object = seurat, method = RPCAIntegration, orig.reduction = "PCA_RNA_split", new.reduction = "integrated.rpca",verbose = FALSE, assay = 'RNA')
seurat <- FindNeighbors(seurat, reduction = "integrated.rpca", dims = 1:30)
seurat <- FindClusters(seurat, resolution = 0.75, cluster.name = "RNA_integrated_clusters_0.75", algorithm = 4)
seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'integrated.rpca', reduction.name = 'UMAP_RNA_integrated')
```

SCT integration workflow: apply SCTransform, split by integration level, perform CCA integration, and save.
```{r integrate-sct, warning=F, message=F }
print(paste('Applying SCTransform and integrating SCT at level:', params$integration_level))
options(future.globals.maxSize = 8000 * 1024^8)

# Create a copy for SCT analysis
seurat_sct <- seurat

# Apply SCT normalization
seurat_sct <- SCTransform(seurat_sct, assay = 'RNA', new.assay.name = 'SCT', verbose = FALSE)

# Split SCT layers by integration group
seurat_sct[["SCT"]] <- split(
  seurat_sct[["SCT"]],
  f = seurat_sct@meta.data[[integration_group_column]]
)

# Perform RPCA integration on SCT assay
seurat_sct <- RunPCA(seurat_sct, assay = 'SCT', reduction.name = 'PCA_SCT_split')
seurat_sct <- IntegrateLayers(object = seurat_sct, method = RPCAIntegration, orig.reduction = "PCA_SCT_split", new.reduction = "integrated.rpca.sct", verbose = FALSE, assay = 'SCT')
seurat_sct <- FindNeighbors(seurat_sct, reduction = "integrated.rpca.sct", dims = 1:30, assay = 'SCT')
seurat_sct <- FindClusters(seurat_sct, resolution = 0.75, cluster.name = "SCT_integrated_clusters_0.75", algorithm = 4)
seurat_sct <- RunUMAP(seurat_sct, dims = 1:30, reduction = 'integrated.rpca.sct', reduction.name = 'UMAP_SCT_integrated', assay = 'SCT')
saveRDS(seurat_sct, file = params$integrated_sct_rds_path)
```

# UMAP plots
## Dim reduction plots
Visual overview: plot integrated UMAP colored by sample_name, batch, and cluster identity.
```{r integrated-umap-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat) <- 'sample_name'
DimPlot(seurat, reduction = 'UMAP_RNA_integrated', raster = F)

Idents(seurat) <- 'batch'
DimPlot(seurat, reduction = 'UMAP_RNA_integrated', raster = F)

Idents(seurat) <- 'RNA_integrated_clusters_0.75'
DimPlot(seurat, reduction = 'UMAP_RNA_integrated', raster = F)
```

## Abundance

Cluster composition: compute and visualize per-sample and per-batch abundance distributions.
```{r cluster-abundance-plots, fig.height=6, fig.width=16, warning=F, message=F, eval = T}
prop_table <- as.data.frame(prop.table(table(seurat$sample_name, seurat$RNA_integrated_clusters_0.75), margin = 1))
sample_batch_map <- unique(seurat@meta.data[, c('sample_name', 'batch')])
sample_batch_map$sample_name <- as.character(sample_batch_map$sample_name)
prop_table$batch <- sample_batch_map$batch[match(as.character(prop_table$Var1), sample_batch_map$sample_name)]
ggplot(prop_table, aes(y = Freq, x = Var2, color = Var1))+geom_point()+theme_classic()
ggplot(prop_table, aes(y = Freq, x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('Frequency')
ggplot(prop_table, aes(y = log10(Freq), x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('log10(Frequency)')
```

## LISI
LISI computation: calculate batch mixing scores from integrated embeddings and save to disk.
```{r compute-lisi, eval = T, warning=F, message=F }
pca <- seurat@reductions$integrated.rpca@cell.embeddings
lisi <- compute_lisi(pca[,c(1:30)], seurat@meta.data, label_colnames = c('batch'))
saveRDS(lisi, file = lisi_rds_path)
```

LISI visualization: plot distributions and spatial patterns of batch mixing scores.
```{r plot-lisi, fig.height=5, fig.width=7, warning=F, message=F, eval = T}
lisi <- readRDS(lisi_rds_path)
seurat$lisi_batch <- lisi$batch
hist(lisi$batch, breaks = 200)
FeaturePlot(seurat, features = 'lisi_batch', raster = F, reduction = 'UMAP_RNA_integrated')
VlnPlot(seurat, features = 'lisi_batch', pt.size = 0)+NoLegend()
```

## QC metrics
QC diagnostics: summarize RNA count/feature, mitochondrial fraction, and Malat1 distributions.
```{r qc-metric-violins, fig.height=6, fig.width=12, warning=F, message=F, eval = T}
VlnPlot(seurat, features = 'nCount_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'nFeature_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'percent.mt', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'Malat1', pt.size = 0)+NoLegend()
```

## Differential expression (cluster markers)
Marker discovery: run bulk-style marker analysis across integrated clusters.
```{r find-cluster-markers, eval = T, warning=F, message=F}
seurat <- JoinLayers(seurat, assay = 'RNA')
FindMarkersBulk(seurat, clus_ident = 'RNA_integrated_clusters_0.75', sample_ident = 'sample_name', n_top_genes = 15, assay = 'RNA', out_dir = 'FindMarkersBulk_RNA_integrated')
```

### Top marker heatmap RNA
Marker heatmap: render scaled average expression heatmap for top RNA markers when available.
```{r marker-heatmap-rna, fig.height=24, fig.width=5, warning=F, message=F, eval = T}
paletteLength <- 50
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
myBreaks <- c(seq(-3, 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(3/paletteLength, 3, length.out=floor(paletteLength/2)))
mat <- NULL
if (!file.exists(markers_csv_path)) {
  message('Skipping RNA marker heatmap because markers file was not found: ', markers_csv_path)
} else {
  markers <- read.csv(markers_csv_path)
  markers <- markers$x
  avg_exp <- AverageExpression(seurat, assays = 'RNA', features = markers, group.by = 'RNA_integrated_clusters_0.75')
  avg_exp <- avg_exp$RNA
  mat <- t(scale(t(avg_exp)))
  pheatmap(mat, color = myColor, breaks = myBreaks, fontsize = 6, cluster_cols = T)
}
```

# SCT Integration Analysis

## SCT UMAP plots
### Dim reduction plots
Visual overview: plot SCT-integrated UMAP colored by sample_name, batch, and cluster identity.
```{r sct-integrated-umap-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat_sct) <- 'sample_name'
DimPlot(seurat_sct, reduction = 'UMAP_SCT_integrated', raster = F)

Idents(seurat_sct) <- 'batch'
DimPlot(seurat_sct, reduction = 'UMAP_SCT_integrated', raster = F)

Idents(seurat_sct) <- 'SCT_integrated_clusters_0.75'
DimPlot(seurat_sct, reduction = 'UMAP_SCT_integrated', raster = F)
```

### Abundance

Cluster composition: compute and visualize per-sample and per-batch abundance distributions for SCT integration.
```{r sct-cluster-abundance-plots, fig.height=6, fig.width=16, warning=F, message=F, eval = T}
prop_table_sct <- as.data.frame(prop.table(table(seurat_sct$sample_name, seurat_sct$SCT_integrated_clusters_0.75), margin = 1))
sample_batch_map <- unique(seurat_sct@meta.data[, c('sample_name', 'batch')])
sample_batch_map$sample_name <- as.character(sample_batch_map$sample_name)
prop_table_sct$batch <- sample_batch_map$batch[match(as.character(prop_table_sct$Var1), sample_batch_map$sample_name)]
ggplot(prop_table_sct, aes(y = Freq, x = Var2, color = Var1))+geom_point()+theme_classic()
ggplot(prop_table_sct, aes(y = Freq, x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('Frequency')
ggplot(prop_table_sct, aes(y = log10(Freq), x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('log10(Frequency)')
```

### LISI
LISI computation: calculate batch mixing scores from SCT-integrated embeddings and save to disk.
```{r compute-lisi-sct, eval = T, warning=F, message=F }
pca_sct <- seurat_sct@reductions$integrated.rpca.sct@cell.embeddings
lisi_sct <- compute_lisi(pca_sct[,c(1:30)], seurat_sct@meta.data, label_colnames = c('batch'))
saveRDS(lisi_sct, file = lisi_sct_rds_path)
```

LISI visualization: plot distributions and spatial patterns of batch mixing scores for SCT integration.
```{r plot-lisi-sct, fig.height=5, fig.width=7, warning=F, message=F, eval = T}
lisi_sct <- readRDS(lisi_sct_rds_path)
seurat_sct$lisi_batch <- lisi_sct$batch
hist(lisi_sct$batch, breaks = 200)
FeaturePlot(seurat_sct, features = 'lisi_batch', raster = F, reduction = 'UMAP_SCT_integrated')
VlnPlot(seurat_sct, features = 'lisi_batch', pt.size = 0)+NoLegend()
```

### QC metrics
QC diagnostics: summarize RNA count/feature, mitochondrial fraction, and Malat1 distributions for SCT integration.
```{r sct-qc-metric-violins, fig.height=6, fig.width=12, warning=F, message=F, eval = T}
VlnPlot(seurat_sct, features = 'nCount_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat_sct, features = 'nFeature_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat_sct, features = 'percent.mt', pt.size = 0)+NoLegend()
VlnPlot(seurat_sct, features = 'Malat1', pt.size = 0)+NoLegend()
```

### Differential expression (cluster markers)
Marker discovery: run bulk-style marker analysis across SCT-integrated clusters.
```{r find-cluster-markers-sct, eval = T, warning=F, message=F}
seurat_sct <- JoinLayers(seurat_sct, assay = 'SCT')
FindMarkersBulk(seurat_sct, clus_ident = 'SCT_integrated_clusters_0.75', sample_ident = 'sample_name', n_top_genes = 15, assay = 'SCT', out_dir = 'FindMarkersBulk_SCT_integrated')
```

### Top marker heatmap SCT
Marker heatmap: render scaled average expression heatmap for top SCT markers when available.
```{r marker-heatmap-sct, fig.height=24, fig.width=5, warning=F, message=F, eval = T}
paletteLength <- 50
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
myBreaks <- c(seq(-3, 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(3/paletteLength, 3, length.out=floor(paletteLength/2)))
mat <- NULL
if (!file.exists(markers_sct_csv_path)) {
  message('Skipping SCT marker heatmap because markers file was not found: ', markers_sct_csv_path)
} else {
  markers <- read.csv(markers_sct_csv_path)
  markers <- markers$x
  avg_exp <- AverageExpression(seurat_sct, assays = 'SCT', features = markers, group.by = 'SCT_integrated_clusters_0.75')
  avg_exp <- avg_exp$SCT
  mat <- t(scale(t(avg_exp)))
  pheatmap(mat, color = myColor, breaks = myBreaks, fontsize = 6, cluster_cols = T)
}
```

