---
title: "Integrate RNA"
author: "Michael Gildea"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: true
    df_print: kable
    code_folding: hide
params:
  merged_rds_path: "merged_QC.rds"
  integrated_rds_path: "integrated.rds"
  integrated_sct_rds_path: "integrated_SCT.rds"
  output_dir: "."
  integration_level: "Batch"
  lisi_rds_path: "lisi_integrated_RNA.rds"
  lisi_sct_rds_path: "lisi_integrated_SCT.rds"
  lisi_harmony_rds_path: "lisi_integrated_RNA_harmony.rds"
  lisi_sct_harmony_rds_path: "lisi_integrated_SCT_harmony.rds"
  markers_csv_path: "FindMarkersBulk_RNA_integrated_rpca/Top_markers.csv"
  markers_sct_csv_path: "FindMarkersBulk_SCT_integrated_rpca/Top_markers.csv"
  markers_harmony_csv_path: "FindMarkersBulk_RNA_integrated_harmony/Top_markers.csv"
  markers_sct_harmony_csv_path: "FindMarkersBulk_SCT_integrated_harmony/Top_markers.csv"
---

<style type="text/css">
  body{
  font-size: 12pt;
}
  h1{
  font-size: 40pt;
}
  h2{
  font-size: 30pt;
}
  h3,h4,h5,h6{
  font-size: 20pt;
}
</style>

Report metadata: print knit timestamp and R version for reproducibility.
```{r report-metadata}
time <- Sys.time()
print(paste('This notebook was knit on:', time))
print(version)
```

Environment context: print working directory used during render.
```{r working-directory}
cat(paste('Bigpurple Location: ',getwd()))
```

Load required R packages for integration, plotting, and diagnostics.
```{r load-libraries, message=F, warning=F}
library(Seurat)
library(ggplot2)
library(pheatmap)
library(lisi)
library(CVRCFunc)
library(harmony)
```

Parameter validation: check integration level and resolve output paths.
```{r validate-params-and-paths, warning=F, message=F}
if (!params$integration_level %in% c('Batch', 'Sample')) {
  stop("params$integration_level must be 'Batch' or 'Sample'.")
}

integration_group_column <- if (params$integration_level == 'Batch') 'batch' else 'sample_name'

resolve_output_path <- function(path_val) {
  if (grepl('^/', path_val)) return(path_val)
  file.path(params$output_dir, path_val)
}

lisi_rds_path <- resolve_output_path(params$lisi_rds_path)
markers_csv_path <- resolve_output_path(params$markers_csv_path)
lisi_sct_rds_path <- resolve_output_path(params$lisi_sct_rds_path)
markers_sct_csv_path <- resolve_output_path(params$markers_sct_csv_path)
lisi_harmony_rds_path <- resolve_output_path(params$lisi_harmony_rds_path)
markers_harmony_csv_path <- resolve_output_path(params$markers_harmony_csv_path)
lisi_sct_harmony_rds_path <- resolve_output_path(params$lisi_sct_harmony_rds_path)
markers_sct_harmony_csv_path <- resolve_output_path(params$markers_sct_harmony_csv_path)
```

Track run progress by logging each code chunk as it starts, with timestamps.
```{r chunk-progress-logging, include=FALSE, warning=F, message=F }
safe_level <- gsub('[^A-Za-z0-9._-]+', '_', params$integration_level)
chunk_log_dir <- file.path(params$output_dir, 'chunk_logs')
dir.create(chunk_log_dir, showWarnings = FALSE, recursive = TRUE)
progress_log <- file.path(chunk_log_dir, paste0('chunk_progress_integrate_RNA_', safe_level, '.log'))
cat(sprintf('\n=== integrate_RNA run started (%s): %s ===\n',
            params$integration_level,
            format(Sys.time(), '%Y-%m-%d %H:%M:%S')),
    file = progress_log,
    append = TRUE)

knitr::opts_hooks$set(label = function(options) {
  chunk_label <- if (is.null(options$label) || !nzchar(options$label)) {
    'unnamed-chunk'
  } else {
    options$label
  }

  line <- sprintf('[%s] START chunk: %s\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), chunk_label)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  options
})

knitr::knit_hooks$set(document = function(x) {
  line <- sprintf('[%s] END render: integrate_RNA (%s)\n',
                  format(Sys.time(), '%Y-%m-%d %H:%M:%S'),
                  params$integration_level)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  x
})
```

Input loading: read merged object and verify required integration metadata exists.
```{r load-merged-object, warning=F, message=F }
if (!file.exists(params$merged_rds_path)) {
  stop('Merged RDS not found: ', params$merged_rds_path)
}
seurat <- readRDS(params$merged_rds_path)

if (!integration_group_column %in% colnames(seurat@meta.data)) {
  stop('Missing integration metadata column: ', integration_group_column)
}
```

Integration workflow: split by selected level, integrate RNA, cluster, and save output RDS.
```{r integrate-log-norm, warning=F, message=F }
print(paste('Integrating RNA at level:', params$integration_level))
options(future.globals.maxSize = 8000 * 1024^8)

seurat[["RNA"]] <- split(
  seurat[["RNA"]],
  f = seurat@meta.data[[integration_group_column]]
)
seurat <- NormalizeData(seurat, assay = 'RNA')
seurat <- FindVariableFeatures(seurat, assay = 'RNA')
seurat <- ScaleData(seurat, assay = 'RNA')
seurat <- RunPCA(seurat, assay = 'RNA', reduction.name = 'PCA_RNA_split')
seurat <- IntegrateLayers(object = seurat, method = RPCAIntegration, orig.reduction = "PCA_RNA_split", new.reduction = "integrated.rpca",verbose = FALSE, assay = 'RNA')
seurat <- FindNeighbors(seurat, reduction = "integrated.rpca", dims = 1:30)
seurat <- FindClusters(seurat, resolution = 0.75, cluster.name = "RNA_integrated_clusters_0.75", algorithm = 4)
seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'integrated.rpca', reduction.name = 'UMAP_RNA_integrated')
```

SCT integration workflow: apply SCTransform, perform rpca integration, and save.
```{r integrate-sct, warning=F, message=F }
print(paste('Applying SCTransform and integrating SCT at level:', params$integration_level))
options(future.globals.maxSize = 8000 * 1024^8)

# Create a copy for SCT analysis
DefaultAssay(seurat) <- "SCT"
seurat <- RunPCA(seurat, assay = 'SCT', reduction.name = 'PCA_SCT_split', row.names(seurat@assays$SCT$scale.data))
seurat <- IntegrateLayers(object = seurat, method = RPCAIntegration, orig.reduction = "PCA_SCT_split", new.reduction = "integrated.rpca.sct",verbose = FALSE, assay = 'SCT', normalization.method = "SCT")
seurat <- FindNeighbors(seurat, reduction = "integrated.rpca.sct", dims = 1:30)
seurat <- FindClusters(seurat, resolution = 0.75, cluster.name = "SCT_integrated_clusters_0.75", algorithm = 4)
seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'integrated.rpca.sct', reduction.name = 'UMAP_SCT_integrated')
```

Harmony workflow (log-normalized RNA): run Harmony on RNA PCA, then cluster and UMAP.
```{r integrate-log-norm-harmony, warning=F, message=F }
print(paste('Running Harmony on RNA at level:', params$integration_level))
DefaultAssay(seurat) <- 'RNA'
seurat <- RunHarmony(
  object = seurat,
  group.by.vars = integration_group_column,
  reduction = 'PCA_RNA_split',
  reduction.save = 'harmony.rna',
  verbose = FALSE
)
seurat <- FindNeighbors(seurat, reduction = 'harmony.rna', dims = 1:30)
seurat <- FindClusters(seurat, resolution = 0.75, cluster.name = 'RNA_harmony_clusters_0.75', algorithm = 4)
seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'harmony.rna', reduction.name = 'UMAP_RNA_harmony')
```

Harmony workflow (SCT): run Harmony on SCT PCA, then cluster and UMAP.
```{r integrate-sct-harmony, warning=F, message=F }
print(paste('Running Harmony on SCT at level:', params$integration_level))
DefaultAssay(seurat) <- 'SCT'
seurat <- RunHarmony(
  object = seurat,
  group.by.vars = integration_group_column,
  reduction = 'PCA_SCT_split',
  reduction.save = 'harmony.sct',
  verbose = FALSE
)
seurat <- FindNeighbors(seurat, reduction = 'harmony.sct', dims = 1:30, assay = 'SCT')
seurat <- FindClusters(seurat, resolution = 0.75, cluster.name = 'SCT_harmony_clusters_0.75', algorithm = 4)
seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'harmony.sct', reduction.name = 'UMAP_SCT_harmony', assay = 'SCT')
```

Save all integration outputs with RPCA and Harmony reductions/clusters.
```{r save-integrated-objects, warning=F, message=F }
saveRDS(seurat, file = params$integrated_rds_path)
saveRDS(seurat, file = params$integrated_sct_rds_path)
```

# LogNormalize RPCA integration analysis
## Dim reduction plots
Visual overview: plot integrated UMAP colored by sample_name, batch, and cluster identity.

### PCA plots
```{r integrated-pca-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat) <- 'sample_name'
DimPlot(seurat, reduction = 'PCA_RNA_split', raster = F)

Idents(seurat) <- 'batch'
DimPlot(seurat, reduction = 'PCA_RNA_split', raster = F)

Idents(seurat) <- 'RNA_integrated_clusters_0.75'
DimPlot(seurat, reduction = 'PCA_RNA_split', raster = F)
```

### UMAP plots
```{r integrated-umap-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat) <- 'sample_name'
DimPlot(seurat, reduction = 'UMAP_RNA_integrated', raster = F)

Idents(seurat) <- 'batch'
DimPlot(seurat, reduction = 'UMAP_RNA_integrated', raster = F)

Idents(seurat) <- 'RNA_integrated_clusters_0.75'
DimPlot(seurat, reduction = 'UMAP_RNA_integrated', raster = F)
```

## Abundance

Cluster composition: compute and visualize per-sample and per-batch abundance distributions.
```{r cluster-abundance-plots, fig.height=6, fig.width=16, warning=F, message=F, eval = T}
prop_table <- as.data.frame(prop.table(table(seurat$sample_name, seurat$RNA_integrated_clusters_0.75), margin = 1))
sample_batch_map <- unique(seurat@meta.data[, c('sample_name', 'batch')])
sample_batch_map$sample_name <- as.character(sample_batch_map$sample_name)
prop_table$batch <- sample_batch_map$batch[match(as.character(prop_table$Var1), sample_batch_map$sample_name)]
ggplot(prop_table, aes(y = Freq, x = Var2, color = Var1))+geom_point()+theme_classic()
ggplot(prop_table, aes(y = Freq, x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('Frequency')
ggplot(prop_table, aes(y = log10(Freq), x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('log10(Frequency)')
```

## LISI
LISI computation: calculate mixing scores for the selected integration level from integrated embeddings and save to disk.
```{r compute-lisi, eval = T, warning=F, message=F }
pca <- seurat@reductions$integrated.rpca@cell.embeddings
lisi <- compute_lisi(pca[,c(1:30)], seurat@meta.data, label_colnames = c(integration_group_column))
saveRDS(lisi, file = lisi_rds_path)
```

LISI visualization: plot distributions and spatial patterns of mixing scores for the selected integration level.
```{r plot-lisi, fig.height=5, fig.width=7, warning=F, message=F, eval = T}
lisi <- readRDS(lisi_rds_path)
lisi_feature_name <- paste0('lisi_', integration_group_column)
seurat[[lisi_feature_name]] <- lisi[[integration_group_column]]
hist(lisi[[integration_group_column]], breaks = 200, main = paste('Integrated RNA LISI -', integration_group_column))
FeaturePlot(seurat, features = lisi_feature_name, raster = F, reduction = 'UMAP_RNA_integrated')
VlnPlot(seurat, features = lisi_feature_name, pt.size = 0)+NoLegend()
```

## QC metrics
QC diagnostics: summarize RNA count/feature, mitochondrial fraction, and Malat1 distributions.
```{r qc-metric-violins, fig.height=6, fig.width=12, warning=F, message=F, eval = T}
VlnPlot(seurat, features = 'nCount_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'nFeature_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'percent.mt', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'Malat1', pt.size = 0)+NoLegend()
```

## Differential expression (cluster markers)
Marker discovery: run bulk-style marker analysis across integrated clusters.
```{r find-cluster-markers, eval = T, warning=F, message=F}
seurat <- JoinLayers(seurat, assay = 'RNA')
FindMarkersBulk(seurat, clus_ident = 'RNA_integrated_clusters_0.75', sample_ident = 'sample_name', n_top_genes = 15, assay = 'RNA', out_dir = 'FindMarkersBulk_RNA_integrated_rpca')
```

### Top marker heatmap RNA
Marker heatmap: render scaled average expression heatmap for top RNA markers when available.
```{r marker-heatmap-rna, fig.height=24, fig.width=5, warning=F, message=F, eval = T}
paletteLength <- 50
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
myBreaks <- c(seq(-3, 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(3/paletteLength, 3, length.out=floor(paletteLength/2)))
mat <- NULL
if (!file.exists(markers_csv_path)) {
  message('Skipping RNA marker heatmap because markers file was not found: ', markers_csv_path)
} else {
  markers <- read.csv(markers_csv_path)
  markers <- markers$x
  avg_exp <- AverageExpression(seurat, assays = 'RNA', features = markers, group.by = 'RNA_integrated_clusters_0.75')
  avg_exp <- avg_exp$RNA
  mat <- t(scale(t(avg_exp)))
  pheatmap(mat, color = myColor, breaks = myBreaks, fontsize = 6, cluster_cols = T)
}
```

# SCT RPCA Integration Analysis

## Dim reduction plots
Visual overview: plot SCT-integrated UMAP colored by sample_name, batch, and cluster identity.

### PCA plots
```{r integrated-sct-pca-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat) <- 'sample_name'
DimPlot(seurat, reduction = 'PCA_SCT_split', raster = F)

Idents(seurat) <- 'batch'
DimPlot(seurat, reduction = 'PCA_SCT_split', raster = F)

Idents(seurat) <- 'SCT_integrated_clusters_0.75'
DimPlot(seurat, reduction = 'PCA_SCT_split', raster = F)
```

```{r sct-integrated-umap-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat) <- 'sample_name'
DimPlot(seurat, reduction = 'UMAP_SCT_integrated', raster = F)

Idents(seurat) <- 'batch'
DimPlot(seurat, reduction = 'UMAP_SCT_integrated', raster = F)

Idents(seurat) <- 'SCT_integrated_clusters_0.75'
DimPlot(seurat, reduction = 'UMAP_SCT_integrated', raster = F)
```

## Abundance

Cluster composition: compute and visualize per-sample and per-batch abundance distributions for SCT integration.
```{r sct-cluster-abundance-plots, fig.height=6, fig.width=16, warning=F, message=F, eval = T}
prop_table_sct <- as.data.frame(prop.table(table(seurat$sample_name, seurat$SCT_integrated_clusters_0.75), margin = 1))
sample_batch_map <- unique(seurat@meta.data[, c('sample_name', 'batch')])
sample_batch_map$sample_name <- as.character(sample_batch_map$sample_name)
prop_table_sct$batch <- sample_batch_map$batch[match(as.character(prop_table_sct$Var1), sample_batch_map$sample_name)]
ggplot(prop_table_sct, aes(y = Freq, x = Var2, color = Var1))+geom_point()+theme_classic()
ggplot(prop_table_sct, aes(y = Freq, x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('Frequency')
ggplot(prop_table_sct, aes(y = log10(Freq), x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('log10(Frequency)')
```

## LISI
LISI computation: calculate mixing scores for the selected integration level from SCT-integrated embeddings and save to disk.
```{r compute-lisi-sct, eval = T, warning=F, message=F }
pca_sct <- seurat@reductions$integrated.rpca.sct@cell.embeddings
lisi_sct <- compute_lisi(pca_sct[,c(1:30)], seurat@meta.data, label_colnames = c(integration_group_column))
saveRDS(lisi_sct, file = lisi_sct_rds_path)
```

LISI visualization: plot distributions and spatial patterns of mixing scores for the selected integration level for SCT integration.
```{r plot-lisi-sct, fig.height=5, fig.width=7, warning=F, message=F, eval = T}
lisi_sct <- readRDS(lisi_sct_rds_path)
lisi_feature_name <- paste0('lisi_', integration_group_column)
seurat[[lisi_feature_name]] <- lisi_sct[[integration_group_column]]
hist(lisi_sct[[integration_group_column]], breaks = 200, main = paste('Integrated SCT LISI -', integration_group_column))
FeaturePlot(seurat, features = lisi_feature_name, raster = F, reduction = 'UMAP_SCT_integrated')
VlnPlot(seurat, features = lisi_feature_name, pt.size = 0)+NoLegend()
```

## QC metrics
QC diagnostics: summarize RNA count/feature, mitochondrial fraction, and Malat1 distributions for SCT integration.
```{r sct-qc-metric-violins, fig.height=6, fig.width=12, warning=F, message=F, eval = T}
VlnPlot(seurat, features = 'nCount_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'nFeature_RNA', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'percent.mt', pt.size = 0)+NoLegend()
VlnPlot(seurat, features = 'Malat1', pt.size = 0)+NoLegend()
```

## Differential expression (cluster markers)
Marker discovery: run bulk-style marker analysis across SCT-integrated clusters.
```{r find-cluster-markers-sct, eval = T, warning=F, message=F}
FindMarkersBulk(seurat, clus_ident = 'SCT_integrated_clusters_0.75', sample_ident = 'sample_name', n_top_genes = 15, assay = 'SCT', out_dir = 'FindMarkersBulk_SCT_integrated_rpca')
```

## Top marker heatmap SCT
Marker heatmap: render scaled average expression heatmap for top SCT markers when available.
```{r marker-heatmap-sct, fig.height=24, fig.width=5, warning=F, message=F, eval = T}
paletteLength <- 50
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
myBreaks <- c(seq(-3, 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(3/paletteLength, 3, length.out=floor(paletteLength/2)))
mat <- NULL
if (!file.exists(markers_sct_csv_path)) {
  message('Skipping SCT marker heatmap because markers file was not found: ', markers_sct_csv_path)
} else {
  markers <- read.csv(markers_sct_csv_path)
  markers <- markers$x
  avg_exp <- AverageExpression(seurat, assays = 'SCT', features = markers, group.by = 'SCT_integrated_clusters_0.75')
  avg_exp <- avg_exp$SCT
  mat <- t(scale(t(avg_exp)))
  pheatmap(mat, color = myColor, breaks = myBreaks, fontsize = 6, cluster_cols = T)
}
```

# LogNormalize Harmony Integration Analysis
## Dim reduction plots
Visual overview: plot Harmony-integrated RNA UMAP colored by sample_name, batch, and cluster identity.

### UMAP plots
```{r harmony-rna-umap-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat) <- 'sample_name'
DimPlot(seurat, reduction = 'UMAP_RNA_harmony', raster = F)

Idents(seurat) <- 'batch'
DimPlot(seurat, reduction = 'UMAP_RNA_harmony', raster = F)

Idents(seurat) <- 'RNA_harmony_clusters_0.75'
DimPlot(seurat, reduction = 'UMAP_RNA_harmony', raster = F)
```

## Abundance

Cluster composition: compute and visualize per-sample and per-batch abundance distributions for RNA Harmony integration.
```{r harmony-rna-cluster-abundance-plots, fig.height=6, fig.width=16, warning=F, message=F, eval = T}
prop_table_harmony <- as.data.frame(prop.table(table(seurat$sample_name, seurat$RNA_harmony_clusters_0.75), margin = 1))
sample_batch_map <- unique(seurat@meta.data[, c('sample_name', 'batch')])
sample_batch_map$sample_name <- as.character(sample_batch_map$sample_name)
prop_table_harmony$batch <- sample_batch_map$batch[match(as.character(prop_table_harmony$Var1), sample_batch_map$sample_name)]
ggplot(prop_table_harmony, aes(y = Freq, x = Var2, color = Var1))+geom_point()+theme_classic()
ggplot(prop_table_harmony, aes(y = Freq, x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('Frequency')
ggplot(prop_table_harmony, aes(y = log10(Freq), x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('log10(Frequency)')
```

## LISI
LISI computation: calculate mixing scores for the selected integration level from Harmony RNA embeddings and save to disk.
```{r compute-lisi-harmony-rna, eval = T, warning=F, message=F }
pca_harmony_rna <- seurat@reductions$harmony.rna@cell.embeddings
lisi_harmony <- compute_lisi(pca_harmony_rna[,c(1:30)], seurat@meta.data, label_colnames = c(integration_group_column))
saveRDS(lisi_harmony, file = lisi_harmony_rds_path)
```

LISI visualization: plot distributions and spatial patterns of mixing scores for RNA Harmony integration.
```{r plot-lisi-harmony-rna, fig.height=5, fig.width=7, warning=F, message=F, eval = T}
lisi_harmony <- readRDS(lisi_harmony_rds_path)
lisi_feature_name_harmony <- paste0('lisi_harmony_rna_', integration_group_column)
seurat[[lisi_feature_name_harmony]] <- lisi_harmony[[integration_group_column]]
hist(lisi_harmony[[integration_group_column]], breaks = 200, main = paste('Harmony RNA LISI -', integration_group_column))
FeaturePlot(seurat, features = lisi_feature_name_harmony, raster = F, reduction = 'UMAP_RNA_harmony')
VlnPlot(seurat, features = lisi_feature_name_harmony, pt.size = 0)+NoLegend()
```

## QC metrics
QC diagnostics: summarize RNA count/feature, mitochondrial fraction, and Malat1 distributions for RNA Harmony integration.
```{r harmony-rna-qc-metric-violins, fig.height=6, fig.width=12, warning=F, message=F, eval = T}
VlnPlot(seurat, features = 'nCount_RNA', pt.size = 0, group.by = 'RNA_harmony_clusters_0.75')+NoLegend()
VlnPlot(seurat, features = 'nFeature_RNA', pt.size = 0, group.by = 'RNA_harmony_clusters_0.75')+NoLegend()
VlnPlot(seurat, features = 'percent.mt', pt.size = 0, group.by = 'RNA_harmony_clusters_0.75')+NoLegend()
VlnPlot(seurat, features = 'Malat1', pt.size = 0, group.by = 'RNA_harmony_clusters_0.75')+NoLegend()
```

## Differential expression (cluster markers)
Marker discovery: run bulk-style marker analysis across RNA Harmony clusters.
```{r find-cluster-markers-harmony-rna, eval = T, warning=F, message=F}
seurat <- JoinLayers(seurat, assay = 'RNA')
FindMarkersBulk(seurat, clus_ident = 'RNA_harmony_clusters_0.75', sample_ident = 'sample_name', n_top_genes = 15, assay = 'RNA', out_dir = 'FindMarkersBulk_RNA_integrated_harmony')
```

## Top marker heatmap RNA Harmony
Marker heatmap: render scaled average expression heatmap for top RNA Harmony markers when available.
```{r marker-heatmap-harmony-rna, fig.height=24, fig.width=5, warning=F, message=F, eval = T}
paletteLength <- 50
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
myBreaks <- c(seq(-3, 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(3/paletteLength, 3, length.out=floor(paletteLength/2)))
if (!file.exists(markers_harmony_csv_path)) {
  message('Skipping RNA Harmony marker heatmap because markers file was not found: ', markers_harmony_csv_path)
} else {
  markers <- read.csv(markers_harmony_csv_path)
  markers <- markers$x
  avg_exp <- AverageExpression(seurat, assays = 'RNA', features = markers, group.by = 'RNA_harmony_clusters_0.75')
  avg_exp <- avg_exp$RNA
  mat <- t(scale(t(avg_exp)))
  pheatmap(mat, color = myColor, breaks = myBreaks, fontsize = 6, cluster_cols = T)
}
```

# SCT Harmony Integration Analysis
## Dim reduction plots
Visual overview: plot SCT Harmony-integrated UMAP colored by sample_name, batch, and cluster identity.
```{r harmony-sct-umap-plots, fig.height=7, fig.width=10, warning=F, message=F, eval = T}

Idents(seurat) <- 'sample_name'
DimPlot(seurat, reduction = 'UMAP_SCT_harmony', raster = F)

Idents(seurat) <- 'batch'
DimPlot(seurat, reduction = 'UMAP_SCT_harmony', raster = F)

Idents(seurat) <- 'SCT_harmony_clusters_0.75'
DimPlot(seurat, reduction = 'UMAP_SCT_harmony', raster = F)
```

## Abundance

Cluster composition: compute and visualize per-sample and per-batch abundance distributions for SCT Harmony integration.
```{r harmony-sct-cluster-abundance-plots, fig.height=6, fig.width=16, warning=F, message=F, eval = T}
prop_table_sct_harmony <- as.data.frame(prop.table(table(seurat$sample_name, seurat$SCT_harmony_clusters_0.75), margin = 1))
sample_batch_map <- unique(seurat@meta.data[, c('sample_name', 'batch')])
sample_batch_map$sample_name <- as.character(sample_batch_map$sample_name)
prop_table_sct_harmony$batch <- sample_batch_map$batch[match(as.character(prop_table_sct_harmony$Var1), sample_batch_map$sample_name)]
ggplot(prop_table_sct_harmony, aes(y = Freq, x = Var2, color = Var1))+geom_point()+theme_classic()
ggplot(prop_table_sct_harmony, aes(y = Freq, x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('Frequency')
ggplot(prop_table_sct_harmony, aes(y = log10(Freq), x = Var2, color = batch))+geom_boxplot()+geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1))+theme_classic()+
  theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1))+
  ylab('log10(Frequency)')
```

## LISI
LISI computation: calculate mixing scores for the selected integration level from Harmony SCT embeddings and save to disk.
```{r compute-lisi-harmony-sct, eval = T, warning=F, message=F }
pca_harmony_sct <- seurat@reductions$harmony.sct@cell.embeddings
lisi_sct_harmony <- compute_lisi(pca_harmony_sct[,c(1:30)], seurat@meta.data, label_colnames = c(integration_group_column))
saveRDS(lisi_sct_harmony, file = lisi_sct_harmony_rds_path)
```

LISI visualization: plot distributions and spatial patterns of mixing scores for SCT Harmony integration.
```{r plot-lisi-harmony-sct, fig.height=5, fig.width=7, warning=F, message=F, eval = T}
lisi_sct_harmony <- readRDS(lisi_sct_harmony_rds_path)
lisi_feature_name_sct_harmony <- paste0('lisi_harmony_sct_', integration_group_column)
seurat[[lisi_feature_name_sct_harmony]] <- lisi_sct_harmony[[integration_group_column]]
hist(lisi_sct_harmony[[integration_group_column]], breaks = 200, main = paste('Harmony SCT LISI -', integration_group_column))
FeaturePlot(seurat, features = lisi_feature_name_sct_harmony, raster = F, reduction = 'UMAP_SCT_harmony')
VlnPlot(seurat, features = lisi_feature_name_sct_harmony, pt.size = 0)+NoLegend()
```

## QC metrics
QC diagnostics: summarize RNA count/feature, mitochondrial fraction, and Malat1 distributions for SCT Harmony integration.
```{r harmony-sct-qc-metric-violins, fig.height=6, fig.width=12, warning=F, message=F, eval = T}
VlnPlot(seurat, features = 'nCount_RNA', pt.size = 0, group.by = 'SCT_harmony_clusters_0.75')+NoLegend()
VlnPlot(seurat, features = 'nFeature_RNA', pt.size = 0, group.by = 'SCT_harmony_clusters_0.75')+NoLegend()
VlnPlot(seurat, features = 'percent.mt', pt.size = 0, group.by = 'SCT_harmony_clusters_0.75')+NoLegend()
VlnPlot(seurat, features = 'Malat1', pt.size = 0, group.by = 'SCT_harmony_clusters_0.75')+NoLegend()
```

## Differential expression (cluster markers)
Marker discovery: run bulk-style marker analysis across SCT Harmony clusters.
```{r find-cluster-markers-harmony-sct, eval = T, warning=F, message=F}
FindMarkersBulk(seurat, clus_ident = 'SCT_harmony_clusters_0.75', sample_ident = 'sample_name', n_top_genes = 15, assay = 'SCT', out_dir = 'FindMarkersBulk_SCT_integrated_harmony')
```

## Top marker heatmap SCT Harmony
Marker heatmap: render scaled average expression heatmap for top SCT Harmony markers when available.
```{r marker-heatmap-harmony-sct, fig.height=24, fig.width=5, warning=F, message=F, eval = T}
paletteLength <- 50
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
myBreaks <- c(seq(-3, 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(3/paletteLength, 3, length.out=floor(paletteLength/2)))
if (!file.exists(markers_sct_harmony_csv_path)) {
  message('Skipping SCT Harmony marker heatmap because markers file was not found: ', markers_sct_harmony_csv_path)
} else {
  markers <- read.csv(markers_sct_harmony_csv_path)
  markers <- markers$x
  avg_exp <- AverageExpression(seurat, assays = 'SCT', features = markers, group.by = 'SCT_harmony_clusters_0.75')
  avg_exp <- avg_exp$SCT
  mat <- t(scale(t(avg_exp)))
  pheatmap(mat, color = myColor, breaks = myBreaks, fontsize = 6, cluster_cols = T)
}
```

# Cross-method LISI comparison
Compare LISI distributions across all four integration methods for the selected integration level.

```{r compare-lisi-across-methods, fig.height=7, fig.width=12, warning=F, message=F, eval = T}
read_lisi_column <- function(path, label_col, method_name) {
  if (!file.exists(path)) {
    message('Skipping ', method_name, ': LISI file not found at ', path)
    return(NULL)
  }

  obj <- readRDS(path)
  if (!(label_col %in% colnames(obj))) {
    message('Skipping ', method_name, ': LISI column ', label_col, ' not found in ', path)
    return(NULL)
  }

  data.frame(
    method = method_name,
    lisi = as.numeric(obj[[label_col]]),
    stringsAsFactors = FALSE
  )
}

lisi_tables <- list(
  read_lisi_column(lisi_rds_path, integration_group_column, 'RNA RPCA'),
  read_lisi_column(lisi_sct_rds_path, integration_group_column, 'SCT RPCA'),
  read_lisi_column(lisi_harmony_rds_path, integration_group_column, 'RNA Harmony'),
  read_lisi_column(lisi_sct_harmony_rds_path, integration_group_column, 'SCT Harmony')
)
lisi_tables <- lisi_tables[!vapply(lisi_tables, is.null, logical(1))]

if (length(lisi_tables) < 2L) {
  message('Need at least two LISI result tables to compare methods. Skipping cross-method LISI comparison.')
} else {
  lisi_compare <- do.call(rbind, lisi_tables)
  lisi_compare <- lisi_compare[is.finite(lisi_compare$lisi), , drop = FALSE]
  lisi_compare$method <- factor(
    lisi_compare$method,
    levels = c('RNA RPCA', 'SCT RPCA', 'RNA Harmony', 'SCT Harmony')
  )
  lisi_compare <- lisi_compare[!is.na(lisi_compare$method), , drop = FALSE]

  if (nrow(lisi_compare) == 0) {
    message('No finite LISI values available for cross-method comparison.')
  } else {
    summary_df <- do.call(rbind, lapply(split(lisi_compare$lisi, lisi_compare$method), function(x) {
      data.frame(
        n_cells = length(x),
        mean = mean(x),
        median = median(x),
        sd = sd(x),
        iqr = IQR(x),
        q05 = as.numeric(quantile(x, 0.05)),
        q95 = as.numeric(quantile(x, 0.95))
      )
    }))
    summary_df$method <- rownames(summary_df)
    rownames(summary_df) <- NULL
    summary_df <- summary_df[, c('method', 'n_cells', 'mean', 'median', 'sd', 'iqr', 'q05', 'q95')]
    print(summary_df)

    print(
      ggplot(lisi_compare, aes(x = method, y = lisi, fill = method)) +
        geom_violin(trim = FALSE, alpha = 0.35, color = NA) +
        geom_boxplot(width = 0.15, outlier.size = 0.15, alpha = 0.8) +
        theme_classic() +
        theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1), legend.position = 'none') +
        xlab('Integration method') +
        ylab(paste('LISI (', integration_group_column, ')', sep = '')) +
        ggtitle(paste('LISI comparison across integration methods -', integration_group_column))
    )

    print(
      ggplot(summary_df, aes(x = method, y = median, fill = method)) +
        geom_col(alpha = 0.8) +
        geom_errorbar(aes(ymin = q05, ymax = q95), width = 0.2) +
        theme_classic() +
        theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1), legend.position = 'none') +
        xlab('Integration method') +
        ylab(paste('Median LISI (', integration_group_column, ') with 5th-95th percentile', sep = '')) +
        ggtitle('Median LISI summary across methods')
    )
  }
}
```

