---
title: "Merge and analyze"
author: "Michael Gildea"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: true
    df_print: kable
    code_folding: hide
---

<style type="text/css">
  body{
  font-size: 12pt;
}
  h1{
  font-size: 40pt;
}
  h2{
  font-size: 30pt;
}
  h3,h4,h5,h6{
  font-size: 20pt;
}
</style>

Print the knit timestamp and R version for run provenance.
```{r session-info}
time <- Sys.time()
print(paste('This notebook was knit on:', time))
print(version)
```

Display the current working directory used during report execution.
```{r working-directory}
cat(paste('Bigpurple Location: ',getwd()))
```

Load all required libraries used by downstream analysis chunks.
```{r load-libraries, message=F, warning=F}
library(Seurat)
library(ggplot2)
library(CVRCFunc)
library(pheatmap)
library(lisi)
library(grid)
```

Initialize chunk-level progress logging so the run records which code block starts executing.
```{r chunk-progress-logging, include=FALSE}
chunk_log_dir <- 'chunk_logs'
dir.create(chunk_log_dir, showWarnings = FALSE, recursive = TRUE)
progress_log <- file.path(normalizePath(chunk_log_dir, mustWork = FALSE), 'chunk_progress_merged.log')
cat(sprintf('\n=== merge_analysis run started: %s ===\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S')),
    file = progress_log, append = TRUE)

knitr::opts_hooks$set(label = function(options) {
  chunk_label <- if (is.null(options$label) || !nzchar(options$label)) {
    'unnamed-chunk'
  } else {
    options$label
  }

  line <- sprintf('[%s] START chunk: %s\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), chunk_label)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  options
})

knitr::knit_hooks$set(chunk = function(x, options) {
  chunk_label <- if (is.null(options$label) || !nzchar(options$label)) {
    'unnamed-chunk'
  } else {
    options$label
  }

  line <- sprintf('[%s] END chunk: %s\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), chunk_label)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  x
})

knitr::knit_hooks$set(document = function(x) {
  line <- sprintf('[%s] END render: merge_analysis\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S'))
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  x
})

template_dir <- normalizePath(knitr::current_input(dir = TRUE), mustWork = FALSE)
qc_support_dir <- file.path(template_dir, 'support_files')

find_qc_support_file <- function(candidates) {
  candidate_paths <- file.path(qc_support_dir, candidates)
  candidate_paths[file.exists(candidate_paths)][1]
}
```

# load merged object
Load the merged Seurat object, harmonize metadata, detect species, and build RNA/SCT reductions if missing.
```{r load-merged-object, warning=F, message=F}
merged_path <- 'merged_QC.rds'
if (!file.exists(merged_path)) {
  stop('Could not find merged object: merged_QC.rds')
}

seurat <- readRDS(merged_path)
message('Loaded merged object: ', merged_path)

DefaultAssay(seurat) <- 'RNA'
seurat <- JoinLayers(seurat, assay = 'RNA')

# Detect species from feature naming conventions so TF annotation can choose
# the correct reference table.
all_features <- rownames(seurat)
human_score <- sum(startsWith(all_features, 'MT-')) +
  sum(all_features %in% c('MALAT1', 'HBA1', 'HBA2', 'HBB'))
mouse_score <- sum(startsWith(all_features, 'mt-')) +
  sum(all_features %in% c('Malat1', 'Hba-a1', 'Hba-a2', 'Hbb-bs', 'Hbb-bt'))
species <- if (human_score >= mouse_score) 'human' else 'mouse'
message('Autodetected species: ', species)

# Harmonize metadata fields expected by this report.
meta_cols <- colnames(seurat@meta.data)
if (!'batch' %in% meta_cols) {
  if ('sample_name' %in% meta_cols) {
    seurat$batch <- seurat$sample_name
  } else if ('orig.ident' %in% meta_cols) {
    seurat$batch <- seurat$orig.ident
  } else {
    seurat$batch <- rep('batch_1', ncol(seurat))
  }
}
if (!'condition' %in% colnames(seurat@meta.data)) {
  seurat$condition <- seurat$batch
}
if (!'sample_name' %in% colnames(seurat@meta.data)) {
  if ('orig.ident' %in% colnames(seurat@meta.data)) {
    seurat$sample_name <- seurat$orig.ident
  } else {
    seurat$sample_name <- seurat$batch
  }
}

# Build merged reductions/clusters if they are not already present.
need_reduction <- !'PCA_RNA_merged' %in% names(seurat@reductions) ||
  !'UMAP_RNA_merged' %in% names(seurat@reductions)
need_clusters <- !'seurat_clusters_merged' %in% colnames(seurat@meta.data)
need_sct_reduction <- !'PCA_SCT_merged' %in% names(seurat@reductions) ||
  !'UMAP_SCT_merged' %in% names(seurat@reductions)
need_sct_clusters <- !'seurat_clusters_sct_merged' %in% colnames(seurat@meta.data)

options(future.globals.maxSize = 8000 * 1024^6) 
if (need_reduction || need_clusters) {
  message('Computing merged normalization/reductions/clusters...')
  set.seed(1234)
  seurat <- NormalizeData(seurat, assay = 'RNA')
  seurat <- FindVariableFeatures(seurat, assay = 'RNA', nfeatures = 3000)
  seurat <- ScaleData(seurat, assay = 'RNA')
  seurat <- RunPCA(seurat, assay = 'RNA', reduction.name = 'PCA_RNA_merged')
  seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'PCA_RNA_merged', assay = 'RNA', reduction.name = 'UMAP_RNA_merged')
  seurat <- FindNeighbors(seurat, assay = 'RNA', reduction = 'PCA_RNA_merged', dims = c(1:30))
  seurat <- FindClusters(seurat, resolution = 0.75, algorithm = 4, cluster.name = 'seurat_clusters_merged')
}

if (need_sct_reduction || need_sct_clusters) {
  message('Computing merged SCTransform reductions/clusters...')
  set.seed(1234)
  DefaultAssay(seurat) <- 'SCT'
  seurat <- RunPCA(seurat, assay = 'SCT', reduction.name = 'PCA_SCT_merged')
  seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'PCA_SCT_merged', assay = 'SCT', reduction.name = 'UMAP_SCT_merged')
  seurat <- FindNeighbors(seurat, assay = 'SCT', reduction = 'PCA_SCT_merged', dims = c(1:30))
  seurat <- FindClusters(seurat, resolution = 0.75, algorithm = 4, cluster.name = 'seurat_clusters_sct_merged')
}

saveRDS(seurat, file = 'merged_QC.rds')
```

# Dim reduction plots
Visualize log-normalization and SCT embeddings colored and split by batch and sample name.
```{r dim-reduction-plots, fig.height=7, fig.width=10, message=FALSE, warning=F}
for (id_col in c('batch', 'sample_name')) {
  if (id_col %in% colnames(seurat@meta.data)) {
    Idents(seurat) <- id_col
    print(DimPlot(seurat, reduction = 'PCA_RNA_merged', raster = F) + ggtitle(paste('RNA PCA by', id_col)))
    print(DimPlot(seurat, reduction = 'UMAP_RNA_merged', raster = F) + ggtitle(paste('RNA UMAP by', id_col)))
    print(DimPlot(seurat, reduction = 'UMAP_RNA_merged', raster = F, split.by = id_col, ncol = 3) + ggtitle(paste('RNA UMAP split by', id_col)))

    print(DimPlot(seurat, reduction = 'PCA_SCT_merged', raster = F) + ggtitle(paste('SCT PCA by', id_col)))
    print(DimPlot(seurat, reduction = 'UMAP_SCT_merged', raster = F) + ggtitle(paste('SCT UMAP by', id_col)))
    print(DimPlot(seurat, reduction = 'UMAP_SCT_merged', raster = F, split.by = id_col, ncol = 3) + ggtitle(paste('SCT UMAP split by', id_col)))
  }
}
print(DimPlot(seurat, reduction = 'UMAP_RNA_merged', group.by = 'seurat_clusters_merged', label = TRUE, raster = F) + ggtitle('log-norm UMAP colored by log-norm clustering'))
print(DimPlot(seurat, reduction = 'UMAP_SCT_merged', group.by = 'seurat_clusters_sct_merged', label = TRUE, raster = F) + ggtitle('SCT UMAP colored by SCT clustering'))
```

# Abundance

Plot per-sample and per-batch cluster abundance distributions for RNA and SCT cluster assignments.
```{r abundance-plots, fig.height=6, fig.width=16, message=F, warning=FALSE}
plot_abundance <- function(cluster_col, method_label) {
  prop_sample <- as.data.frame(prop.table(table(seurat$sample_name, seurat[[cluster_col]][, 1]), margin = 1))
  sample_batch <- unique(seurat@meta.data[, c('sample_name', 'batch')])
  prop_sample$sample_name <- prop_sample$Var1
  prop_sample_batch <- merge(prop_sample, sample_batch, by = 'sample_name', all.x = TRUE, sort = FALSE)

  list(
    sample_plot = ggplot(prop_sample, aes(y = Freq, x = Var2, color = Var1)) +
      geom_boxplot() +
      geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1)) +
      theme_classic() +
      theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1)) +
      ylab('Frequency') +
      ggtitle(paste(method_label, 'abundance by sample_name')) +
      labs(color = 'sample_name'),
    batch_plot = ggplot(prop_sample_batch, aes(y = Freq, x = Var2, color = batch, fill = batch)) +
      geom_boxplot(aes(group = interaction(Var2, batch)), alpha = 0.15, outlier.shape = NA,
                   position = position_dodge(width = .8)) +
      geom_point(position = position_jitterdodge(dodge.width = .7, jitter.width = .1), size = 1.8) +
      theme_classic() +
      theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1)) +
      ylab('Frequency') +
      ggtitle(paste(method_label, 'abundance by batch')) +
      labs(color = 'batch', fill = 'batch')
  )
}

rna_abundance_plots <- plot_abundance('seurat_clusters_merged', 'log-norm')
print(rna_abundance_plots$sample_plot)
print(rna_abundance_plots$batch_plot)

sct_abundance_plots <- plot_abundance('seurat_clusters_sct_merged', 'SCT')
print(sct_abundance_plots$sample_plot)
print(sct_abundance_plots$batch_plot)
```

# LISI
Compute LISI (Local Inverse Simpson's Index) scores from RNA and SCT PCA spaces and save them for reuse.
LISI quantifies local neighborhood diversity for metadata labels (here: batch and sample_name) around each cell.
Higher LISI means stronger local mixing across label groups (typically less batch/sample separation), while lower LISI means neighborhoods are dominated by fewer labels (stronger separation).
In practice, consistently low batch-LISI with clear batch-segregated embeddings suggests technical batch effects and supports trying integration, while moderate/high batch-LISI with biologically coherent clustering suggests integration may be unnecessary.
```{r compute-lisi, warning=F, message=F}
pca_rna <- seurat@reductions$PCA_RNA_merged@cell.embeddings
pca_sct <- seurat@reductions$PCA_SCT_merged@cell.embeddings
lisi <- list(
  RNA = compute_lisi(pca_rna[, c(1:30)], seurat@meta.data, label_colnames = c('batch', 'sample_name')),
  SCT = compute_lisi(pca_sct[, c(1:30)], seurat@meta.data, label_colnames = c('batch', 'sample_name'))
)
saveRDS(lisi, file = 'lisi.rds')
```

Load or recompute LISI, then visualize distributions and embedding-level patterns for both methods.
Histograms show the global spread of mixing scores; FeaturePlot overlays reveal where mixing is high or low in UMAP space; violin plots summarize per-cell score distributions for each label type.
Use these plots to compare RNA vs SCT normalization in terms of local batch/sample mixing behavior.
```{r lisi-visualizations, fig.height=5, fig.width=7, warning=F, message=F}
if (file.exists('lisi.rds')) {
  lisi <- readRDS('lisi.rds')
} else {
  pca_rna <- seurat@reductions$PCA_RNA_merged@cell.embeddings
  pca_sct <- seurat@reductions$PCA_SCT_merged@cell.embeddings
  lisi <- list(
    RNA = compute_lisi(pca_rna[, c(1:30)], seurat@meta.data, label_colnames = c('batch', 'sample_name')),
    SCT = compute_lisi(pca_sct[, c(1:30)], seurat@meta.data, label_colnames = c('batch', 'sample_name'))
  )
  saveRDS(lisi, file = 'lisi.rds')
}

for (method in c('RNA', 'SCT')) {
  lisi_df <- lisi[[method]]
  reduction_name <- if (method == 'RNA') 'UMAP_RNA_merged' else 'UMAP_SCT_merged'
  for (lisi_col in c('batch', 'sample_name')) {
    if (lisi_col %in% colnames(lisi_df)) {
      feature_name <- paste0('lisi_', tolower(method), '_', lisi_col)
      seurat[[feature_name]] <- lisi_df[[lisi_col]]
      hist(lisi_df[[lisi_col]], breaks = 200, main = paste(method, 'LISI -', lisi_col))
      print(FeaturePlot(seurat, features = feature_name, reduction = reduction_name, raster = F))
      print(VlnPlot(seurat, features = feature_name, pt.size = 0) + NoLegend())
    }
  }
}
```

# QC metrics
Compare key QC metrics across log-normalization and SCT-based cluster identities.

## log-norm clusters
```{r qc-lognorm-violin, fig.height=10, fig.width=20, warning=F, message=F}
qc_features <- c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'S.Score', 'G2M.Score')

VlnPlot(seurat, features = qc_features,
        group.by = 'seurat_clusters_merged', pt.size = 0) + NoLegend()
```

```{r qc-lognorm-featureplots, fig.height=5, fig.width=7, warning=F, message=F}
for (feature_name in qc_features) {
  if (!(feature_name %in% colnames(seurat@meta.data))) {
    message('Skipping missing QC metric: ', feature_name)
    next
  }

  print(
    FeaturePlot(seurat,
                features = feature_name,
                reduction = 'UMAP_RNA_merged',
                raster = F) +
      ggtitle(paste('log-norm UMAP -', feature_name))
  )
}
```

## SCT clusters
```{r qc-sct-violin, fig.height=10, fig.width=20, warning=F, message=F}
qc_features <- c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'S.Score', 'G2M.Score')

VlnPlot(seurat, features = qc_features,
        group.by = 'seurat_clusters_sct_merged', pt.size = 0) + NoLegend()
```

```{r qc-sct-featureplots, fig.height=5, fig.width=7, warning=F, message=F}
for (feature_name in qc_features) {
  if (!(feature_name %in% colnames(seurat@meta.data))) {
    message('Skipping missing QC metric: ', feature_name)
    next
  }

  print(
    FeaturePlot(seurat,
                features = feature_name,
                reduction = 'UMAP_SCT_merged',
                raster = F) +
      ggtitle(paste('SCT UMAP -', feature_name))
  )
}
```


# Differential expression (cluster markers)
Run pseudobulk-style cluster marker discovery for log-norm and SCT cluster labels.
```{r differential-expression-markers, warning=F, message=F}
FindMarkersBulk(seurat, clus_ident = 'seurat_clusters_merged', sample_ident = 'sample_name', n_top_genes = 15, assay = 'RNA', out_dir = 'FindMarkersBulk_merged_log_norm')
FindMarkersBulk(seurat, clus_ident = 'seurat_clusters_sct_merged', sample_ident = 'sample_name', n_top_genes = 15, assay = 'RNA', out_dir = 'FindMarkersBulk_merged_SCT')
```

## Top marker heatmap log-norm
Build heatmaps of top markers from log-norm and SCT marker result tables.
```{r top-marker-heatmap, fig.height=30, fig.width=7, warning=F, message=F}
plot_marker_heatmap <- function(marker_csv, group_col, assay_name, plot_title) {
  if (!file.exists(marker_csv)) {
    message('Skipping marker heatmap: ', marker_csv, ' not found.')
    return(invisible(NULL))
  }

  paletteLength <- 50
  myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
  myBreaks <- c(seq(-3, 0, length.out=ceiling(paletteLength/2) + 1), 
                seq(3/paletteLength, 3, length.out=floor(paletteLength/2)))
  markers <- read.csv(marker_csv)
  markers <- markers$x
  avg_exp <- AverageExpression(seurat, assays = assay_name, features = markers, group.by = group_col)
  avg_exp <- avg_exp[[assay_name]]
  mat <- t(scale(t(avg_exp)))
  pheatmap(mat, color = myColor, breaks = myBreaks, fontsize = 6, cluster_cols = T, main = plot_title)
  invisible(mat)
}

mat_rna <- plot_marker_heatmap('FindMarkersBulk_merged_log_norm/Top_markers.csv', 'seurat_clusters_merged', 'RNA', 'Top marker heatmap - log-norm')
mat_sct <- plot_marker_heatmap('FindMarkersBulk_merged_SCT/Top_markers.csv', 'seurat_clusters_sct_merged', 'SCT', 'Top marker heatmap - SCT')
```

Annotate marker heatmaps by labeling transcription factor genes when reference files are available.
```{r tf-labeled-heatmap, fig.height=10, fig.width=7, warning=F, message=F}
add_tf_labels <- function(mat, label_text) {
  if (is.null(mat)) {
    message('Skipping TF-labeled heatmap: marker matrix not available for ', label_text)
    return(invisible(NULL))
  }

  tf_candidates <- if (species == 'human') {
    c('allTFs_hg38.txt', 'allTFs_mm.txt')
  } else {
    c('allTFs_mm.txt', 'allTFs_hg38.txt')
  }
  tf_path <- find_qc_support_file(tf_candidates)

  genes <- row.names(mat)
  if (!is.na(tf_path)) {
    TFs <- read.table(tf_path, header = F, sep = '\t')
    genes[which(!row.names(mat) %in% TFs$V1)] <- ''
  } else {
    genes[] <- ''
  }

  paletteLength <- 50
  myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
  myBreaks <- c(seq(-2, 0, length.out=ceiling(paletteLength/2) + 1), 
                seq(max(mat)/paletteLength, 2, length.out=floor(paletteLength/2)))

  add.flag <- function(pheatmap,
                       kept.labels,
                       repel.degree) {

  # repel.degree = number within [0, 1], which controls how much 
  #                space to allocate for repelling labels.
  ## repel.degree = 0: spread out labels over existing range of kept labels
  ## repel.degree = 1: spread out labels over the full y-axis

  heatmap <- pheatmap$gtable

  new.label <- heatmap$grobs[[which(heatmap$layout$name == "row_names")]] 

  kept.labels <- unique(kept.labels[nzchar(kept.labels)])
  if (!length(kept.labels)) {
    grid.newpage()
    grid.draw(heatmap)
    return(invisible(heatmap))
  }

  # keep only labels in kept.labels, replace the rest with ""
  new.label$label <- ifelse(new.label$label %in% kept.labels, 
                            new.label$label, "")

  if (!any(new.label$label != "")) {
    grid.newpage()
    grid.draw(heatmap)
    return(invisible(heatmap))
  }

  # calculate evenly spaced out y-axis positions
  repelled.y <- function(d, d.select, k = repel.degree){
    # d = vector of distances for labels
    # d.select = vector of T/F for which labels are significant

    # recursive function to get current label positions
    # (note the unit is "npc" for all components of each distance)
    strip.npc <- function(dd){
      if(!"unit.arithmetic" %in% class(dd)) {
        return(as.numeric(dd))
      }

      d1 <- strip.npc(dd$arg1)
      d2 <- strip.npc(dd$arg2)
      fn <- dd$fname
      return(lazyeval::lazy_eval(paste(d1, fn, d2)))
    }

    full.range <- sapply(seq_along(d), function(i) strip.npc(d[i]))
    selected.range <- sapply(seq_along(d[d.select]), function(i) strip.npc(d[d.select][i]))

    return(unit(seq(from = max(selected.range) + k*(max(full.range) - max(selected.range)),
                    to = min(selected.range) - k*(min(selected.range) - min(full.range)), 
                    length.out = sum(d.select)), 
                "npc"))
  }
  new.y.positions <- repelled.y(new.label$y,
                                d.select = new.label$label != "")
  new.flag <- segmentsGrob(x0 = new.label$x,
                           x1 = new.label$x + unit(0.15, "npc"),
                           y0 = new.label$y[new.label$label != ""],
                           y1 = new.y.positions)

  # shift position for selected labels
  new.label$x <- new.label$x + unit(0.2, "npc")
  new.label$y[new.label$label != ""] <- new.y.positions

  # add flag to heatmap
  heatmap <- gtable::gtable_add_grob(x = heatmap,
                                   grobs = new.flag,
                                   t = 4, 
                                   l = 4
  )

  # replace label positions in heatmap
  heatmap$grobs[[which(heatmap$layout$name == "row_names")]] <- new.label

  # plot result
  grid.newpage()
  grid.draw(heatmap)

  # return a copy of the heatmap invisibly
  invisible(heatmap)
}

  heat <- pheatmap(
    mat,
    cluster_rows = T,
    cluster_cols = T,
    color = myColor,
    breaks = myBreaks,
    main = paste('Top 15 overexpressed genes per cluster\nScaled average expression\nTranscription factors labeled -', label_text),
    silent = T,
    fontsize = 8
  )
  add.flag(heat, kept.labels = genes, repel.degree = 0.5)
}

add_tf_labels(mat_rna, 'log-norm')
add_tf_labels(mat_sct, 'SCT')
```

# Doublets
Visualize merged doublet annotations on both RNA and SCT UMAPs and summarize doublet-class composition across log-norm and SCT clusters.

```{r doublet-umap-highlights, fig.height=7, fig.width=10, warning=F, message=F}
doublet_runs <- data.frame(
  class_col = c('doublet_class_clustered', 'doublet_class_unclustered', 'doublet_class'),
  run_label = c('clustered', 'unclustered', 'legacy/default'),
  stringsAsFactors = FALSE
)
doublet_runs <- doublet_runs[doublet_runs$class_col %in% colnames(seurat@meta.data), , drop = FALSE]

if (nrow(doublet_runs) == 0) {
  message('Skipping doublet highlight UMAPs: no doublet class metadata columns found.')
} else {
  for (i in seq_len(nrow(doublet_runs))) {
    class_col <- doublet_runs$class_col[i]
    run_label <- doublet_runs$run_label[i]

    doublet_cells <- rownames(seurat@meta.data)[tolower(as.character(seurat[[class_col]][, 1])) == 'doublet']

    if (!length(doublet_cells)) {
      message('Skipping ', run_label, ' doublet highlight UMAPs: no cells labeled as doublet were found in ', class_col, '.')
      next
    }

    print(
      DimPlot(seurat,
              reduction = 'UMAP_RNA_merged',
              cells.highlight = doublet_cells,
              cols = 'grey80',
              cols.highlight = 'firebrick',
              sizes.highlight = 0.6,
              raster = F) +
        ggtitle(paste('log-norm UMAP highlighting doublets (', run_label, ')', sep = ''))
    )

    print(
      DimPlot(seurat,
              reduction = 'UMAP_SCT_merged',
              cells.highlight = doublet_cells,
              cols = 'grey80',
              cols.highlight = 'firebrick',
              sizes.highlight = 0.6,
              raster = F) +
        ggtitle(paste('SCT UMAP highlighting doublets (', run_label, ')', sep = ''))
    )
  }
}
```

```{r doublet-frequency-by-cluster, fig.height=6, fig.width=14, warning=F, message=F}
plot_doublet_frequency <- function(cluster_col, reduction_label, class_col, run_label) {
  required_cols <- c(cluster_col, class_col)
  if (!all(required_cols %in% colnames(seurat@meta.data))) {
    message('Skipping ', reduction_label, ' ', run_label, ' doublet frequency plot: missing metadata columns.')
    return(invisible(NULL))
  }

  doublet_freq <- as.data.frame(prop.table(table(seurat[[cluster_col]][, 1], seurat[[class_col]][, 1]), margin = 1))
  colnames(doublet_freq) <- c('cluster', 'doublet_class', 'frequency')

  print(
    ggplot(doublet_freq, aes(x = cluster, y = frequency, fill = doublet_class)) +
      geom_col(position = 'stack') +
      theme_classic() +
      theme(axis.text.x = element_text(angle = 315, hjust = 0, vjust = 1)) +
      ylab('Fraction of cells') +
      xlab('Cluster') +
      ggtitle(paste('Doublet class frequency by', reduction_label, 'clusters (', run_label, ')', sep = ''))
  )

  invisible(doublet_freq)
}

doublet_runs <- data.frame(
  class_col = c('doublet_class_clustered', 'doublet_class_unclustered', 'doublet_class'),
  run_label = c('clustered', 'unclustered', 'legacy/default'),
  stringsAsFactors = FALSE
)
doublet_runs <- doublet_runs[doublet_runs$class_col %in% colnames(seurat@meta.data), , drop = FALSE]

if (nrow(doublet_runs) == 0) {
  message('Skipping doublet frequency plots: no doublet class metadata columns found.')
} else {
  for (i in seq_len(nrow(doublet_runs))) {
    plot_doublet_frequency(
      cluster_col = 'seurat_clusters_merged',
      reduction_label = 'log-norm',
      class_col = doublet_runs$class_col[i],
      run_label = doublet_runs$run_label[i]
    )

    plot_doublet_frequency(
      cluster_col = 'seurat_clusters_sct_merged',
      reduction_label = 'SCT',
      class_col = doublet_runs$class_col[i],
      run_label = doublet_runs$run_label[i]
    )
  }
}
```
