---
title: "`r paste(params$sample_name, 'QC')`"
author: "Michael Gildea"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: true
    df_print: kable
    code_folding: hide
params:
  sample_name:      "Sample"
  batch:            "A"
  filtered_path:    ""
  raw_path:         ""
  use_cellbender:   FALSE
  support_dir:      "support_files"
  output_dir:       "."
  # Per-metric hard thresholds. Leave as NA to use the MAD-based defaults.
  min_nCount_RNA:   NA
  max_nCount_RNA:   NA
  min_nFeature_RNA: NA
  max_nFeature_RNA: NA
  max_percent_mt:   NA
  min_malat1:       0
  min_nCount_ADT:   NA
  max_nCount_ADT:   NA
  # Hashtag (HTO) demultiplexing and CITE-seq (ADT) options. Leave use_hashtag /
  # use_adt FALSE to skip these sections entirely (default, matches prior behavior).
  use_hashtag:            FALSE
  use_adt:                FALSE
  feature_reference_path: ""
  hto_path:               ""
  adt_path:               ""
  raw_adt_path:           ""
  hto_features:           ""
  dsb_background_rna_max:  NA
  dsb_background_prot_min: NA
  dsb_background_prot_max: NA
---

<style type="text/css">
  body { font-size: 12pt; }
  h1   { font-size: 30pt; }
  h2   { font-size: 20pt; }
  h3, h4, h5, h6 { font-size: 15pt; }
</style>

```{r setup, include=FALSE}
# Set global chunk defaults: suppress messages and warnings throughout
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

template_dir <- normalizePath(knitr::current_input(dir = TRUE), mustWork = FALSE)
qc_support_dir <- normalizePath(params$support_dir, mustWork = FALSE)

find_qc_support_file <- function(filename) {
  candidates <- unique(c(
    file.path(qc_support_dir, filename),
    file.path(template_dir, 'support_files', filename),
    file.path(getwd(), 'support_files', filename)
  ))

  found <- candidates[file.exists(candidates)]
  if (length(found) > 0L) return(found[[1L]])

  stop(
    'Support file not found: ', filename,
    '\nSearched:\n  ', paste(candidates, collapse = '\n  '),
    call. = FALSE
  )
}

source(find_qc_support_file('qc_helpers.R'))
```

Track run progress by logging each code block as it starts, with timestamps.
```{r chunk-progress-logging, include=FALSE}
safe_sample_name <- gsub('[^A-Za-z0-9._-]+', '_', params$sample_name)
chunk_log_dir <- file.path(params$output_dir, 'chunk_logs')
dir.create(chunk_log_dir, showWarnings = FALSE, recursive = TRUE)
progress_log <- file.path(chunk_log_dir, paste0('chunk_progress_', safe_sample_name, '.log'))
cat(sprintf('\n=== sample_QC run started for %s: %s ===\n',
            params$sample_name,
            format(Sys.time(), '%Y-%m-%d %H:%M:%S')),
    file = progress_log,
    append = TRUE)

param_lines <- c(
  sprintf('[%s] PARAM sample_name=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$sample_name)),
  sprintf('[%s] PARAM batch=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$batch)),
  sprintf('[%s] PARAM filtered_path=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$filtered_path)),
  sprintf('[%s] PARAM raw_path=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$raw_path)),
  sprintf('[%s] PARAM output_dir=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$output_dir)),
  sprintf('[%s] PARAM min_nCount_RNA=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$min_nCount_RNA)),
  sprintf('[%s] PARAM max_nCount_RNA=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$max_nCount_RNA)),
  sprintf('[%s] PARAM min_nFeature_RNA=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$min_nFeature_RNA)),
  sprintf('[%s] PARAM max_nFeature_RNA=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$max_nFeature_RNA)),
  sprintf('[%s] PARAM max_percent_mt=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$max_percent_mt)),
  sprintf('[%s] PARAM min_malat1=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$min_malat1)),
  sprintf('[%s] PARAM min_nCount_ADT=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$min_nCount_ADT)),
  sprintf('[%s] PARAM max_nCount_ADT=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$max_nCount_ADT)),
  sprintf('[%s] PARAM use_hashtag=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$use_hashtag)),
  sprintf('[%s] PARAM use_adt=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$use_adt)),
  sprintf('[%s] PARAM feature_reference_path=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$feature_reference_path)),
  sprintf('[%s] PARAM hto_path=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$hto_path)),
  sprintf('[%s] PARAM adt_path=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$adt_path)),
  sprintf('[%s] PARAM raw_adt_path=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$raw_adt_path)),
  sprintf('[%s] PARAM hto_features=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$hto_features)),
  sprintf('[%s] PARAM dsb_background_rna_max=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$dsb_background_rna_max)),
  sprintf('[%s] PARAM dsb_background_prot_min=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$dsb_background_prot_min)),
  sprintf('[%s] PARAM dsb_background_prot_max=%s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), as.character(params$dsb_background_prot_max))
)
cat(paste0(param_lines, '\n'), file = progress_log, append = TRUE)

knitr::opts_hooks$set(label = function(options) {
  chunk_label <- if (is.null(options$label) || !nzchar(options$label)) {
    'unnamed-chunk'
  } else {
    options$label
  }

  line <- sprintf('[%s] START chunk: %s\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), chunk_label)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  options
})

knitr::knit_hooks$set(document = function(x) {
  line <- sprintf('[%s] END render: %s\n', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), params$sample_name)
  cat(line, file = progress_log, append = TRUE)
  message(sub('\n$', '', line))
  x
})
```

```{r session-info}
# Record knit time and R version for reproducibility
time <- Sys.time()
print(paste('This notebook was knit on:', time))
print(version)
```

```{r working-directory}
# Confirm working directory on the cluster
cat(paste('Working directory:', getwd()))
```

```{r load-libraries}
# ── Visualisation ──────────────────────────────────────────────────────────────
library(ggplot2)
library(patchwork)   # combine ggplot2 panels with +
library(viridis)
library(ggrepel)

# ── Data manipulation ──────────────────────────────────────────────────────────
library(dplyr)
library(Hmisc)
library(knitr)

# ── Single-cell analysis ───────────────────────────────────────────────────────
library(Seurat)
library(scDblFinder)
library(presto)


```


# Read in data and plot QC metrics
Loads the CellRanger-filtered matrix for QC analysis and the unfiltered (raw) matrix for barcode rank plotting. Species (human/mouse) is auto-detected from gene symbol conventions and used to configure all downstream metric patterns.

```{r load-data}
# Load either 10x MEX directories or 10x/CellBender .h5 files.
# This pipeline assumes the sample sheet points directly at each matrix input.
# Read10X()/Read10X_h5() return either a single matrix (RNA-only input) or a
# named list keyed by feature type (e.g. "Gene Expression", "Antibody Capture",
# "Multiplexing Capture") for CellRanger multi/feature-barcoded matrices.
read_10x_input_list <- function(input_path, label) {
  if (dir.exists(input_path)) {
    return(Read10X(input_path))
  }

  if (file.exists(input_path) && grepl("\\.(h5|hdf5)$", input_path, ignore.case = TRUE)) {
    return(Read10X_h5(input_path))
  }

  stop(
    "Could not read ", label, " input: ", input_path,
    ". Expected a 10x MEX directory or an .h5/.hdf5 file.",
    call. = FALSE
  )
}

# Extract a named feature-type matrix from Read10X()/Read10X_h5() output.
# Returns NULL if the requested feature type is not present.
extract_feature_type <- function(x, feature_type, label, input_path) {
  if (inherits(x, "dgCMatrix") || is.matrix(x)) {
    if (identical(feature_type, "Gene Expression")) return(x)
    return(NULL)
  }

  if (!is.list(x) || length(x) == 0L) {
    stop("Unsupported ", label, " input format at: ", input_path, call. = FALSE)
  }

  if (is.null(names(x)) || !any(nzchar(names(x)))) {
    if (length(x) == 1L) {
      return(if (identical(feature_type, "Gene Expression")) x[[1L]] else NULL)
    }
    stop(
      "Multiple unnamed matrices detected in ", label, " input: ", input_path,
      ". Expected named feature types (e.g. 'Gene Expression', 'Antibody Capture').",
      call. = FALSE
    )
  }

  match_idx <- which(tolower(names(x)) == tolower(feature_type))
  if (length(match_idx) == 0L) return(NULL)
  x[[match_idx[1L]]]
}

read_10x_input <- function(input_path, label) {
  rna <- extract_feature_type(read_10x_input_list(input_path, label), "Gene Expression", label, input_path)
  if (is.null(rna)) {
    stop("Could not find a 'Gene Expression' matrix in ", label, " input: ", input_path, call. = FALSE)
  }
  rna
}

read_10x_assay_input <- function(input_path, label, feature_types) {
  select_10x_assay_matrix(
    read_10x_input_list(input_path, label),
    feature_types = feature_types,
    label = label,
    input_path = input_path
  )
}

# Resolve HTO/ADT matrices for a given 10x input, in priority order:
# explicit hto_path/adt_path > native "Multiplexing Capture" (HTO only) >
# split_antibody_capture() on "Antibody Capture".
resolve_hto_adt <- function(input_path, label, want_hto, want_adt,
                             hto_path, adt_path, feature_names, hto_features) {
  result <- list(hto = NULL, adt = NULL)
  if (!want_hto && !want_adt) return(result)

  if (want_hto && nzchar(trimws(hto_path))) {
    result$hto <- read_10x_assay_input(
      hto_path, paste0(label, ' HTO'),
      feature_types = c('Multiplexing Capture', 'Antibody Capture')
    )
  }
  if (want_adt && nzchar(trimws(adt_path))) {
    result$adt <- read_10x_assay_input(
      adt_path, paste0(label, ' ADT'),
      feature_types = 'Antibody Capture'
    )
  }

  still_need_hto <- want_hto && is.null(result$hto)
  still_need_adt <- want_adt && is.null(result$adt)
  if (still_need_hto || still_need_adt) {
    raw_list <- read_10x_input_list(input_path, label)
    if (still_need_hto) {
      mux <- extract_feature_type(raw_list, 'Multiplexing Capture', label, input_path)
      if (!is.null(mux)) result$hto <- mux
    }
    ab <- extract_feature_type(raw_list, 'Antibody Capture', label, input_path)
    if (!is.null(ab) && ((still_need_hto && is.null(result$hto)) || still_need_adt)) {
      split_res <- split_antibody_capture(ab, feature_names, hto_features)
      if (still_need_hto && is.null(result$hto)) result$hto <- split_res$hto
      if (still_need_adt) result$adt <- split_res$adt
    }
  }

  if (want_hto && is.null(result$hto)) {
    stop(
      'use_hashtag TRUE but no HTO features could be resolved from ', label, ' input: ', input_path,
      '. Provide hto_path or ensure hto_features matches feature_reference.csv names.',
      call. = FALSE
    )
  }
  if (want_adt && is.null(result$adt)) {
    stop(
      'use_adt TRUE but no ADT features could be resolved from ', label, ' input: ', input_path,
      '. Provide adt_path or ensure feature_reference.csv names include non-HTO Antibody Capture features.',
      call. = FALSE
    )
  }

  result
}

if (isTRUE(params$use_cellbender)) {
  if (!file.exists(params$filtered_path) || !grepl("\\.(h5|hdf5)$", params$filtered_path, ignore.case = TRUE)) {
    stop(
      "--use_cellbender TRUE requires filtered_path to be a direct .h5 or .hdf5 file path (for example, cellbender_filtered.h5). Got: ",
      params$filtered_path,
      call. = FALSE
    )
  }
}

describe_loaded_files <- function(input_path) {
  if (dir.exists(input_path)) {
    children <- list.files(input_path, full.names = TRUE, recursive = FALSE)
    files <- children[file.exists(children)]
    keep <- files[grepl("(matrix\\.mtx(\\.gz)?|barcodes\\.tsv(\\.gz)?|(features|genes)\\.tsv(\\.gz)?)$",
                       basename(files), ignore.case = TRUE)]
    if (length(keep) > 0L) return(sort(normalizePath(keep, mustWork = FALSE)))
    return(normalizePath(input_path, mustWork = FALSE))
  }

  if (file.exists(input_path)) {
    return(normalizePath(input_path, mustWork = TRUE))
  }

  input_path
}

feature_reference_names <- NULL
if (isTRUE(params$use_hashtag) || isTRUE(params$use_adt)) {
  if (is.na(params$feature_reference_path) ||
      !nzchar(trimws(params$feature_reference_path)) ||
      !file.exists(params$feature_reference_path)) {
    stop(
      'feature_reference_path must point to an existing feature_reference.csv when use_hashtag or use_adt is TRUE for sample ',
      params$sample_name, '.',
      call. = FALSE
    )
  }
  feature_reference_names <- read_feature_reference_names(params$feature_reference_path)
}
validate_raw_adt_path(params$use_adt, params$raw_adt_path)

seurat     <- read_10x_input(params$filtered_path, "filtered")
raw_matrix <- read_10x_input(params$raw_path, "raw")

# Resolve optional HTO (cell hashing) and ADT (CITE-seq) matrices; no-op when
# neither use_hashtag nor use_adt is set (default), leaving RNA-only behavior unchanged.
hto_matrix_filtered <- NULL
adt_matrix_filtered <- NULL
adt_matrix_raw      <- NULL

if (isTRUE(params$use_hashtag) || isTRUE(params$use_adt)) {
  filtered_hto_adt <- resolve_hto_adt(
    input_path = params$filtered_path, label = 'filtered',
    want_hto = isTRUE(params$use_hashtag), want_adt = isTRUE(params$use_adt),
    hto_path = params$hto_path, adt_path = params$adt_path,
    feature_names = feature_reference_names, hto_features = params$hto_features
  )
  hto_matrix_filtered <- filtered_hto_adt$hto
  adt_matrix_filtered <- filtered_hto_adt$adt

  if (isTRUE(params$use_adt)) {
    # DSB background counts always come from the explicit raw ADT input.
    raw_hto_adt <- resolve_hto_adt(
      input_path = params$raw_path, label = 'raw',
      want_hto = FALSE, want_adt = TRUE,
      hto_path = '', adt_path = params$raw_adt_path,
      feature_names = feature_reference_names, hto_features = params$hto_features
    )
    adt_matrix_raw <- raw_hto_adt$adt
  }

  cat('HTO features resolved:', if (!is.null(hto_matrix_filtered)) nrow(hto_matrix_filtered) else 0, '\n')
  cat('ADT features resolved:', if (!is.null(adt_matrix_filtered)) nrow(adt_matrix_filtered) else 0, '\n')
}

filtered_loaded <- describe_loaded_files(params$filtered_path)
raw_loaded <- describe_loaded_files(params$raw_path)

cat('Filtered input path:', params$filtered_path, '\n')
cat('Raw input path:', params$raw_path, '\n')
cat('Filtered file(s) loaded:\n')
cat(paste0('  - ', filtered_loaded), sep = '\n')
cat('\nRaw file(s) loaded:\n')
cat(paste0('  - ', raw_loaded), sep = '\n')
cat('\n')

if (exists('progress_log')) {
  log_lines <- c(
    sprintf('[%s] INPUT filtered path: %s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), params$filtered_path),
    if (length(filtered_loaded) > 0L) sprintf('[%s] INPUT filtered file: %s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), filtered_loaded) else character(0),
    sprintf('[%s] INPUT raw path: %s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), params$raw_path),
    if (length(raw_loaded) > 0L) sprintf('[%s] INPUT raw file: %s', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), raw_loaded) else character(0)
  )
  cat(paste0(log_lines, '\n'), file = progress_log, append = TRUE)
}

# Detect species from feature naming conventions so QC metrics adapt to
# human/mouse gene symbol casing and marker patterns.
all_features <- rownames(seurat)
human_score <- sum(startsWith(all_features, 'MT-')) +
  sum(all_features %in% c('MALAT1', 'HBA1', 'HBA2', 'HBB'))
mouse_score <- sum(startsWith(all_features, 'mt-')) +
  sum(all_features %in% c('Malat1', 'Hba-a1', 'Hba-a2', 'Hbb-bs', 'Hbb-bt'))

species <- if (human_score >= mouse_score) 'human' else 'mouse'
mt_pattern <- if (species == 'human') '^MT-' else '^mt-'

malat1_candidates <- if (species == 'human') c('MALAT1', 'Malat1') else c('Malat1', 'MALAT1')
malat1_feature <- malat1_candidates[malat1_candidates %in% all_features][1]

# Load species-specific hemoglobin gene lists from CSVs located alongside the template.
# Each CSV has one gene symbol per line, no header.
hb_csv         <- find_qc_support_file(
                    if (species == 'human') 'human_hemoglobin_genes.csv'
                    else                    'mouse_hemoglobin_genes.csv')
hb_gene_list   <- readLines(hb_csv)
hb_gene_list   <- trimws(hb_gene_list[nzchar(hb_gene_list)])
rbc_features   <- intersect(hb_gene_list, all_features)

cat('Autodetected species:', species, '\n')
cat('Mitochondrial gene pattern:', mt_pattern, '\n')
if (!is.na(malat1_feature)) {
  cat('MALAT1 feature used:', malat1_feature, '\n')
} else {
  cat('MALAT1 feature not found; MALAT1 filter will be skipped.\n')
}
cat('RBC genes detected:', length(rbc_features), '\n')
```

# Barcode rank plots
Horizontal red line marks the UMI threshold determined by CellRanger that separates putative cells from empty droplets. If cellbender was used, the horizontal line corresponds to the minimum non-zero UMI barcode.
The second panel shows a moving local slope of the log-rank vs log-UMI curve and labels an ambient-peak candidate in the empty-droplet tail. This is still in development.
If `--use_cellbender TRUE` was used, an additional barcode rank panel is generated from the CellBender filtered matrix.

```{r barcode-rank-plot, fig.width=11, fig.height=9}
# Rank all barcodes by descending UMI count and plot the knee curve;
# the CellRanger cell/empty threshold is shown as a horizontal red line
# and a moving local slope is used to highlight the ambient RNA shelf
bc_rank_data          <- data.frame(UMI_counts = as.numeric(Matrix::colSums(raw_matrix)))
bc_rank_data          <- arrange(bc_rank_data, desc(UMI_counts))
bc_rank_data$rank     <- seq_len(nrow(bc_rank_data))
filtered_col_sums     <- as.numeric(Matrix::colSums(seurat))
positive_filtered     <- filtered_col_sums[is.finite(filtered_col_sums) & filtered_col_sums > 0]
if (length(positive_filtered) > 0) {
  cellranger_count_thr <- min(positive_filtered)
  cellranger_threshold <- log10(cellranger_count_thr)
} else {
  cellranger_count_thr <- NA_real_
  cellranger_threshold <- NA_real_
  cat('No positive filtered-barcode UMI totals were found; skipping CellRanger threshold lines.\n')
}

bc_rank_data$log_rank <- log10(bc_rank_data$rank)
bc_rank_data$log_umi  <- log10(pmax(bc_rank_data$UMI_counts, 1))
x_limits <- range(bc_rank_data$log_rank, finite = TRUE)

# Rolling local slope d(log10(UMI))/d(log10(rank)) using a centered window.
# Use a vectorized centered finite difference rather than fitting one linear
# model per barcode, which is slow on large raw matrices.
n_bc <- nrow(bc_rank_data)
slope_window <- max(101, as.integer(0.01 * n_bc))
if (slope_window %% 2 == 0) slope_window <- slope_window + 1
slope_window <- min(slope_window, n_bc - (1 - n_bc %% 2))

moving_slope <- rep(NA_real_, n_bc)
half_w <- (slope_window - 1) / 2
if (n_bc >= slope_window) {
  center_idx <- seq_len(n_bc)
  left_idx <- pmax(center_idx - half_w, 1)
  right_idx <- pmin(center_idx + half_w, n_bc)
  valid_span <- right_idx > left_idx
  moving_slope[valid_span] <-
    (bc_rank_data$log_umi[right_idx[valid_span]] - bc_rank_data$log_umi[left_idx[valid_span]]) /
    (bc_rank_data$log_rank[right_idx[valid_span]] - bc_rank_data$log_rank[left_idx[valid_span]])
}
bc_rank_data$moving_slope <- moving_slope

if (is.finite(cellranger_count_thr)) {
  threshold_candidates <- which(bc_rank_data$UMI_counts >= cellranger_count_thr)
  threshold_rank <- if (length(threshold_candidates) > 0) max(threshold_candidates) else NA_real_
  threshold_rank_log <- if (is.finite(threshold_rank) && threshold_rank > 0) log10(threshold_rank) else NA_real_
} else {
  threshold_rank <- NA_real_
  threshold_rank_log <- NA_real_
}

valid_slope <- bc_rank_data %>%
  filter(UMI_counts > 0,
         !is.na(moving_slope)) %>%
  arrange(rank)

ambient_peak <- NULL
first_drop_rank <- NA_real_
second_drop_rank <- NA_real_
ambient_shelf <- NULL
average_ambient_rna_umi <- NA_real_

# Ambient shelf rule: moving from low rank to high rank, find the first two
# local minima in slope and identify the shelf between them.
if (nrow(valid_slope) > 20) {
  slope_vals <- valid_slope$moving_slope
  if (length(slope_vals) >= 7) {
    smooth_k <- max(11L, as.integer(0.005 * length(slope_vals)))
    if (smooth_k %% 2 == 0) smooth_k <- smooth_k + 1L
    smooth_k <- min(smooth_k, length(slope_vals) - (1L - length(slope_vals) %% 2L))
    if (smooth_k >= 3) {
      slope_smooth <- as.numeric(stats::filter(
        slope_vals,
        filter = rep(1 / smooth_k, smooth_k),
        sides = 2,
        method = 'convolution'
      ))
    } else {
      slope_smooth <- slope_vals
    }
  } else {
    slope_smooth <- slope_vals
  }

  finite_idx <- which(is.finite(slope_smooth))
  if (length(finite_idx) >= 3) {
    slope_finite <- slope_smooth[finite_idx]
    local_min_finite_idx <- which(
      slope_finite[2:(length(slope_finite) - 1)] <= slope_finite[1:(length(slope_finite) - 2)] &
      slope_finite[2:(length(slope_finite) - 1)] <= slope_finite[3:length(slope_finite)]
    ) + 1

    if (length(local_min_finite_idx) >= 2) {
      local_min_idx <- finite_idx[local_min_finite_idx]
      deep_cutoff <- as.numeric(quantile(slope_finite, probs = 0.20, names = FALSE))
      deep_min_idx <- local_min_idx[slope_smooth[local_min_idx] <= deep_cutoff]

      selected_min_idx <- if (length(deep_min_idx) >= 2) {
        deep_min_idx
      } else {
        local_min_idx[order(slope_smooth[local_min_idx])[seq_len(min(2, length(local_min_idx)))]]
      }

      if (length(selected_min_idx) >= 2) {
        selected_min_idx <- sort(selected_min_idx)
        first_drop_rank <- valid_slope$rank[selected_min_idx[1]]
        second_drop_rank <- valid_slope$rank[selected_min_idx[2]]

        shelf_candidates <- valid_slope[
          valid_slope$rank > first_drop_rank & valid_slope$rank < second_drop_rank,
          , drop = FALSE
        ]

        if (nrow(shelf_candidates) > 0) {
          ambient_shelf <- shelf_candidates
          average_ambient_rna_umi <- mean(ambient_shelf$UMI_counts, na.rm = TRUE)
          best_idx <- which.min(abs(shelf_candidates$moving_slope))
          ambient_peak <- shelf_candidates[best_idx, , drop = FALSE]
        }
      }
    }
  }
}

# Fallback: if shelf boundaries are not found, choose the closest-to-zero slope
# in the empty-droplet region when available.
if (is.null(ambient_peak) && nrow(valid_slope) > 0) {
  fallback_set <- valid_slope
  if (is.finite(cellranger_count_thr)) {
    fallback_set <- valid_slope[valid_slope$UMI_counts <= cellranger_count_thr, , drop = FALSE]
  }
  if (nrow(fallback_set) > 0) {
    ambient_shelf <- fallback_set
    average_ambient_rna_umi <- mean(ambient_shelf$UMI_counts, na.rm = TRUE)
    best_idx <- which.min(abs(fallback_set$moving_slope))
    ambient_peak <- fallback_set[best_idx, , drop = FALSE]
  }
}

if (is.finite(average_ambient_rna_umi)) {
  cat('Average ambient RNA UMI (ambient shelf mean):', signif(average_ambient_rna_umi, 4),
      'across', nrow(ambient_shelf), 'barcodes\n')
}

first_drop_rank_log <- if (is.finite(first_drop_rank) && first_drop_rank > 0) log10(first_drop_rank) else NA_real_
second_drop_rank_log <- if (is.finite(second_drop_rank) && second_drop_rank > 0) log10(second_drop_rank) else NA_real_
slope_limits <- range(bc_rank_data$moving_slope, finite = TRUE)
slope_span <- diff(slope_limits)
if (!is.finite(slope_span) || slope_span <= 0) slope_span <- 1

rank_plot <- ggplot(bc_rank_data, aes(x = log_rank, y = log_umi)) +
  geom_point(color = 'navy') +
  ggtitle('Barcode rank plot – RNA') +
  theme_classic() +
  scale_x_continuous(limits = x_limits) +
  ylab('log10(UMI counts)') +
  xlab('log10(barcode rank)')

if (is.finite(cellranger_threshold)) {
  rank_plot <- rank_plot +
    geom_hline(yintercept = cellranger_threshold, color = 'firebrick', linetype = 'dashed')
}

slope_plot <- ggplot(bc_rank_data, aes(x = log_rank, y = moving_slope)) +
  geom_line(color = 'darkgreen', linewidth = 0.4, na.rm = TRUE) +
  theme_classic() +
  ggtitle('Moving local slope of barcode rank curve') +
  labs(subtitle = 'Ambient shelf: between first two deep local minima of smoothed slope') +
  scale_x_continuous(limits = x_limits) +
  ylab('Local slope') +
  xlab('log10(barcode rank)')

if (is.finite(threshold_rank_log)) {
  slope_plot <- slope_plot +
    geom_vline(xintercept = threshold_rank_log, color = 'firebrick', linetype = 'dashed')
}

if (is.finite(first_drop_rank_log)) {
  slope_plot <- slope_plot +
    geom_vline(xintercept = first_drop_rank_log, color = 'steelblue4', linetype = 'dotted')
}

if (is.finite(second_drop_rank_log)) {
  slope_plot <- slope_plot +
    geom_vline(xintercept = second_drop_rank_log, color = 'steelblue4', linetype = 'dotted')
}

if (!is.null(ambient_peak) && nrow(ambient_peak) == 1) {
  ambient_label_y <- ambient_peak$moving_slope - (0.12 * slope_span)
  ambient_label_y <- max(slope_limits[1] + (0.08 * slope_span), ambient_label_y)
  ambient_label_y <- min(slope_limits[2] - (0.10 * slope_span), ambient_label_y)

  slope_plot <- slope_plot +
    geom_point(data = ambient_peak,
               aes(x = log_rank, y = moving_slope),
               color = 'orange3', size = 2.2, inherit.aes = FALSE) +
    geom_vline(xintercept = ambient_peak$log_rank,
               color = 'orange3', linetype = 'dotdash') +
    annotate('label',
             x = ambient_peak$log_rank,
             y = ambient_label_y,
             label = paste0('ambient shelf candidate\nrank=', ambient_peak$rank,
                            ', UMI=', ambient_peak$UMI_counts,
                            '\navg ambient shelf UMI=', signif(average_ambient_rna_umi, 4)),
             hjust = 1.05,
             size = 3,
             label.size = 0.2,
             fill = 'white')
}

rank_plot / slope_plot + plot_layout(heights = c(4, 4))

# Show a second rank plot only when --use_cellbender TRUE was requested.
is_cellbender_input <- isTRUE(params$use_cellbender)
if (is_cellbender_input) {
  cb_rank_data <- data.frame(UMI_counts = as.numeric(Matrix::colSums(seurat)))
  cb_rank_data <- arrange(cb_rank_data, desc(UMI_counts))
  cb_rank_data$rank <- seq_len(nrow(cb_rank_data))
  cb_rank_data$log_rank <- log10(cb_rank_data$rank)
  cb_rank_data$log_umi <- log10(pmax(cb_rank_data$UMI_counts, 1))

  cellbender_rank_plot <- ggplot(cb_rank_data, aes(x = log_rank, y = log_umi)) +
    geom_point(color = 'purple4', alpha = 0.6, size = 0.7) +
    theme_classic() +
    ggtitle('Barcode rank plot – filtered input (CellBender)') +
    xlab('log10(barcode rank)') +
    ylab('log10(UMI counts)')

  if (is.finite(cellranger_threshold)) {
    cellbender_rank_plot <- cellbender_rank_plot +
      geom_hline(yintercept = cellranger_threshold, color = 'firebrick')
  }

  cellbender_rank_plot
} else {
  cat('Skipping CellBender barcode rank panel (--use_cellbender TRUE was not set).\n')
}
```

# CellBender vs raw gene-level UMI comparison
Compares total per-gene UMIs in the CellBender filtered matrix versus the raw matrix, restricted to barcodes retained in the CellBender matrix. Only the top 25% most changed genes by absolute percent difference are plotted, and the top 30 are labeled. Genes with a total UMI count below 10 in either matrix are excluded.

```{r cellbender-vs-raw-gene-umi-scatter, fig.width=9, fig.height=7}
if (!isTRUE(params$use_cellbender)) {
  cat('Skipping CellBender-vs-raw gene UMI scatter (--use_cellbender TRUE was not set).\n')
} else {
  common_barcodes <- intersect(colnames(seurat), colnames(raw_matrix))
  common_genes <- intersect(rownames(seurat), rownames(raw_matrix))

  if (length(common_barcodes) == 0L || length(common_genes) == 0L) {
    cat('Skipping CellBender-vs-raw gene UMI scatter: no overlapping barcodes or genes.\n')
  } else {
    cb_gene_totals <- as.numeric(Matrix::rowSums(seurat[common_genes, common_barcodes, drop = FALSE]))
    raw_gene_totals <- as.numeric(Matrix::rowSums(raw_matrix[common_genes, common_barcodes, drop = FALSE]))

    gene_umi_cmp <- data.frame(
      gene = common_genes,
      raw_umi = raw_gene_totals,
      cellbender_umi = cb_gene_totals,
      stringsAsFactors = FALSE
    )

    gene_umi_cmp <- gene_umi_cmp[
      gene_umi_cmp$raw_umi >= 10 & gene_umi_cmp$cellbender_umi >= 10,
      , drop = FALSE
    ]

    if (nrow(gene_umi_cmp) == 0L) {
      cat('Skipping CellBender-vs-raw gene UMI scatter: no genes had >= 10 UMI in both matrices.\n')
    } else {

      gene_umi_cmp$pct_diff <- 100 * (gene_umi_cmp$cellbender_umi - gene_umi_cmp$raw_umi) / gene_umi_cmp$raw_umi
      gene_umi_cmp$abs_pct_diff <- abs(gene_umi_cmp$pct_diff)

      finite_abs_diff <- gene_umi_cmp$abs_pct_diff[is.finite(gene_umi_cmp$abs_pct_diff)]
      if (length(finite_abs_diff) == 0L) {
        cat('Skipping CellBender-vs-raw gene UMI scatter: percent differences are non-finite for all genes.\n')
      } else {
        diff_cutoff <- as.numeric(quantile(finite_abs_diff, probs = 0.75, names = FALSE, na.rm = TRUE))
        genes_to_plot <- gene_umi_cmp[!is.finite(gene_umi_cmp$abs_pct_diff) |
                                        gene_umi_cmp$abs_pct_diff >= diff_cutoff, , drop = FALSE]

        if (nrow(genes_to_plot) == 0L) {
          cat('Skipping CellBender-vs-raw gene UMI scatter: no genes passed top-25% change filter.\n')
        } else {
          label_n <- min(30L, nrow(genes_to_plot))
          top_label_genes <- genes_to_plot[order(genes_to_plot$abs_pct_diff, decreasing = TRUE), , drop = FALSE][seq_len(label_n), , drop = FALSE]

          cat('Plotting', nrow(genes_to_plot), 'genes (top 25% most changed by absolute percent difference) across',
              length(common_barcodes), 'CellBender-retained barcodes.\n')

          ggplot(genes_to_plot, aes(x = raw_umi + 1, y = cellbender_umi + 1)) +
            geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = 'grey50') +
            geom_point(aes(color = abs_pct_diff), alpha = 0.7, size = 1.2) +
            ggrepel::geom_text_repel(
              data = top_label_genes,
              aes(label = gene),
              size = 2.5,
              color = 'black',
              max.overlaps = Inf,
              force = 1,
              box.padding = 0.25,
              point.padding = 0.1,
              min.segment.length = 0
            ) +
            scale_x_log10() +
            scale_y_log10() +
            scale_color_viridis_c(option = 'magma', name = 'abs % diff') +
            theme_classic() +
            ggtitle('Per-gene total UMI: CellBender vs raw (CellBender barcodes only)') +
            xlab('Raw total UMI per gene (+1, log10)') +
            ylab('CellBender total UMI per gene (+1, log10)')
        }
      }
    }
  }
}
```

# Filter for features present in at least 0.1% of cells
Removes genes detected in fewer than 0.1% of barcodes before Seurat object creation to reduce matrix sparsity and speed up downstream steps without meaningfully affecting QC.

```{r filter-lowly-detected-genes}
# Remove genes detected in fewer than 0.1% of barcodes to reduce matrix sparsity
min_cells  <- 0.001 * ncol(seurat)
cat('Keeping genes detected in at least', min_cells, 'cells\n')
keep_genes <- Matrix::rowSums(seurat) >= min_cells
seurat     <- seurat[keep_genes, ]
```

# Create Seurat object
Wraps the sparse count matrix in a Seurat object and computes the three primary per-cell QC covariates: UMI count (`nCount_RNA`), genes detected (`nFeature_RNA`), and mitochondrial read fraction (`percent.mt`, pattern chosen by species).

```{r create-seurat-object}
# Require at least 10 UMIs per cell before creating the Seurat object.
cell_umi_totals <- Matrix::colSums(seurat)
keep_cells_10 <- cell_umi_totals >= 10
cat('Keeping', sum(keep_cells_10), 'of', length(keep_cells_10), 'cells with >= 10 UMIs\n')
seurat <- seurat[, keep_cells_10, drop = FALSE]

# Convert the filtered sparse matrix to a Seurat object and compute per-cell QC metrics
seurat               <- CreateSeuratObject(seurat, min.cells = 0, min.features = 0)
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, pattern = mt_pattern)
seurat$Sample        <- rep(params$sample_name, ncol(seurat))
seurat$batch         <- rep(params$batch, ncol(seurat))
```

```{r attach-hto-adt-assays}
# Attach optional HTO/ADT assays, restricting to barcodes present in all
# requested assays. No-op (seurat unchanged) when neither assay was resolved.
if (!is.null(hto_matrix_filtered) || !is.null(adt_matrix_filtered)) {
  common_cells <- colnames(seurat)
  if (!is.null(hto_matrix_filtered)) common_cells <- intersect(common_cells, colnames(hto_matrix_filtered))
  if (!is.null(adt_matrix_filtered)) common_cells <- intersect(common_cells, colnames(adt_matrix_filtered))

  dropped_n <- ncol(seurat) - length(common_cells)
  if (dropped_n > 0) {
    cat('Dropping', dropped_n, 'RNA cells missing HTO/ADT barcodes before attaching HTO/ADT assays.\n')
    seurat <- seurat[, common_cells]
  }

  if (!is.null(hto_matrix_filtered)) {
    seurat[['HTO']] <- CreateAssayObject(counts = hto_matrix_filtered[, common_cells, drop = FALSE])
    cat('Attached HTO assay with', nrow(hto_matrix_filtered), 'features.\n')
  }
  if (!is.null(adt_matrix_filtered)) {
    seurat[['ADT']] <- CreateAssayObject(counts = adt_matrix_filtered[, common_cells, drop = FALSE])
    cat('Attached ADT assay with', nrow(adt_matrix_filtered), 'features.\n')
  }
}
```

```{r hto-section-header, results='asis', echo=FALSE}
if (isTRUE(params$use_hashtag)) {
  cat(
    '# Hashtag demultiplexing\n\n',
    'Demultiplexes pooled hashtagged cells into per-cell sample identity using ',
    '`HTODemux`, with diagnostic plots and an intra-sample doublet check via ',
    '`scDblFinder::recoverDoublets`.\n\n',
    sep = ''
  )
}
```

```{r htodemux, eval=isTRUE(params$use_hashtag)}
# HTODemux requires CLR-normalized HTO counts (independent of ADT's DSB choice).
seurat <- NormalizeData(seurat, assay = 'HTO', normalization.method = 'CLR', margin = 2)
seurat <- HTODemux(seurat, assay = 'HTO', positive.quantile = 0.99)

seurat$HTO_sample   <- seurat$hash.ID
seurat$HTO_class    <- seurat$HTO_classification.global
seurat$replicate_id <- paste(params$sample_name, seurat$hash.ID, sep = '_')

cat('HTO classification counts:\n')
print(table(seurat$HTO_class))
```

```{r hto-qc-plots1, eval=isTRUE(params$use_hashtag), fig.height=10, fig.width=18}
HTOHeatmap(seurat, assay = 'HTO')
RidgePlot(seurat, assay = 'HTO', features = rownames(seurat[['HTO']]), ncol = 2)
```

```{r hto-qc-plots2, eval=isTRUE(params$use_hashtag), fig.height=8, fig.width=10}
# Per-HTO CLR density plots show assigned cells against all other cells.
VlnPlot(seurat, features = 'nCount_RNA', group.by = 'HTO_class', log = TRUE, pt.size = 0.1)
hto_data <- GetAssayData(seurat, assay = 'HTO', layer = 'data')
hto_thresholds <- hto_singlet_expression_thresholds(
  hto_data,
  hto_class = seurat$HTO_class,
  hash_id = seurat$hash.ID
)
for (hto_feature in rownames(hto_data)) {
  gg_data <- data.frame(HTO_counts = hto_data[hto_feature, ], hash_ID = as.character(seurat$hash.ID))
  gg_data$hash_ID[gg_data$hash_ID != hto_feature] <- 'Other'
  p <- ggplot(gg_data, aes(x = HTO_counts, fill = hash_ID, color = hash_ID)) +
    geom_density(alpha = 0.5) +
    theme_classic() +
    ggtitle(hto_feature)
  print(p)

  all_cell_expression <- data.frame(HTO_expression = as.numeric(hto_data[hto_feature, ]))
  hto_threshold <- unname(hto_thresholds[[hto_feature]])
  all_cell_histogram <- ggplot(all_cell_expression, aes(x = HTO_expression)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50,
                   fill = 'grey80', color = 'white') +
    geom_density(color = 'steelblue4', linewidth = 0.8, na.rm = TRUE) +
    theme_classic() +
    labs(
      title = paste(hto_feature, 'CLR expression across all cells'),
      subtitle = if (is.finite(hto_threshold)) {
        paste('Minimum Singlet expression:', signif(hto_threshold, 4))
      } else {
        'No Singlet cells assigned to this HTO'
      },
      x = 'CLR-normalized HTO expression',
      y = 'Density'
    )
  if (is.finite(hto_threshold)) {
    all_cell_histogram <- all_cell_histogram +
      geom_vline(xintercept = hto_threshold, color = 'red', linewidth = 0.8)
  }
  print(all_cell_histogram)
}
```

```{r hto-remove-negatives, eval=isTRUE(params$use_hashtag)}
# Negatives lack a hashtag call and are always excluded (unusable identity).
n_before_neg <- ncol(seurat)
seurat <- subset(seurat, subset = HTO_class != 'Negative')
cat('Removed', n_before_neg - ncol(seurat), 'HTO Negative cells;', ncol(seurat), 'remain (Singlet + Doublet).\n')
```

```{r hto-intra-sample-doublets, eval=isTRUE(params$use_hashtag), fig.height=5, fig.width=12}
# Intra-sample doublets: same-hashtag doublets HTODemux can't catch, detected
# via scDblFinder::recoverDoublets on an RNA PCA/UMAP restricted to this subset.
if (!requireNamespace('scDblFinder', quietly = TRUE) || !requireNamespace('scater', quietly = TRUE) ||
    !requireNamespace('gridExtra', quietly = TRUE)) {
  cat('scDblFinder/scater/gridExtra not installed; skipping intra-sample doublet check.\n')
  seurat$predicted_intra_sample_doublet <- FALSE
} else {
  hto_check <- seurat
  DefaultAssay(hto_check) <- 'RNA'
  hto_check <- NormalizeData(hto_check, assay = 'RNA', verbose = FALSE)
  hto_check <- FindVariableFeatures(hto_check, assay = 'RNA', verbose = FALSE)
  hto_check <- ScaleData(hto_check, assay = 'RNA', verbose = FALSE)
  hto_check <- RunPCA(hto_check, assay = 'RNA', npcs = 30, verbose = FALSE, reduction.name = 'pca_hto_check')
  hto_check <- RunUMAP(hto_check, reduction = 'pca_hto_check', dims = 1:30, verbose = FALSE, reduction.name = 'umap_hto_check')

  hto_sce <- as.SingleCellExperiment(hto_check)
  hto_dimreds <- resolve_hto_doublet_dimreds(SingleCellExperiment::reducedDimNames(hto_sce))
  doublets_pred <- scDblFinder::recoverDoublets(
    hto_sce,
    use.dimred = hto_dimreds$pca,
    doublets = (hto_sce$HTO_class == 'Doublet'),
    samples = table(hto_sce$hash.ID)
  )
  hto_sce$predicted_intra_sample_doublet <- doublets_pred$predicted

  seurat$predicted_intra_sample_doublet <- FALSE
  seurat$predicted_intra_sample_doublet[colnames(hto_sce)] <- hto_sce$predicted_intra_sample_doublet

  cat('Predicted intra-sample doublets:', sum(hto_sce$predicted_intra_sample_doublet), '/', ncol(hto_sce), '\n')

  gridExtra::grid.arrange(
    scater::plotUMAP(hto_sce, colour_by = 'HTO_class', dimred = hto_dimreds$umap) + ggtitle('Known HTO classification'),
    scater::plotUMAP(hto_sce, colour_by = 'predicted_intra_sample_doublet', dimred = hto_dimreds$umap) + ggtitle('Predicted intra-sample doublets'),
    ncol = 2
  )
}
```

```{r hto-filter-singlets, eval=isTRUE(params$use_hashtag)}
n_before <- ncol(seurat)
keep_singlets <- hto_singlet_keep_mask(seurat$HTO_class)
seurat <- subset(seurat, cells = colnames(seurat)[keep_singlets])
cat('Kept', ncol(seurat), 'of', n_before, 'cells classified as HTO Singlet (removed known HTO Doublets).\n')
```

# Cell and gene count prior to filtering
Prints the Seurat object summary to record the number of cells and features entering the QC filtering step.

```{r pre-filter-object-summary}
seurat
```

```{r hto-cell-numbers, eval=isTRUE(params$use_hashtag)}
table(seurat$hash.ID)
```

# Summary statistics
Descriptive statistics for the primary per-cell QC metrics; ADT UMI statistics are included when ADT is enabled.

## UMI count
Per-cell UMI count distribution. Very low values indicate empty droplets; very high values may indicate multiplets.
```{r summary-umi}
print(describe(seurat$nCount_RNA))
```

## Genes detected
Number of unique genes detected per cell. Complements UMI count for identifying outlier cells at both extremes.
```{r summary-genes}
print(describe(seurat$nFeature_RNA))
```

## Fraction mitochondrial alignments
Proportion of reads mapping to mitochondrial genes. Elevated values indicate stressed, lysed, or low-quality cells.
```{r summary-mito}
print(describe(seurat$percent.mt))
```

```{r summary-adt-count, eval=isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat)}
print(describe(seurat$nCount_ADT))
```

# Median absolute deviation (MAD)-based filtering
Use the tables below to choose appropriate cutoff multiples before applying the filter.
Thresholds are the metric value corresponding to ±1–5 MADs from the sample median.

## Compute MAD thresholds
Calculates per-metric medians and MADs, then tabulates the metric value at ±1–5 MAD multiples together with the fraction and count of cells that fall on each side. Use these tables to select appropriate filter multiples.

```{r compute-mad-thresholds}
# ── Per-metric medians and MADs ───────────────────────────────────────────────
nCount_RNA_med   <- median(seurat$nCount_RNA, na.rm = TRUE)
nCount_RNA_mad   <- median(abs(seurat$nCount_RNA   - nCount_RNA_med), na.rm = TRUE)

nFeature_RNA_med <- median(seurat$nFeature_RNA, na.rm = TRUE)
nFeature_RNA_mad <- median(abs(seurat$nFeature_RNA - nFeature_RNA_med), na.rm = TRUE)

percent.mt_med   <- median(seurat$percent.mt, na.rm = TRUE)
percent.mt_mad   <- median(abs(seurat$percent.mt   - percent.mt_med), na.rm = TRUE)

cat('nCount_RNA   median =', nCount_RNA_med,   '| MAD =', nCount_RNA_mad,   '\n')
cat('nFeature_RNA median =', nFeature_RNA_med, '| MAD =', nFeature_RNA_mad, '\n')
cat('percent.mt   median =', percent.mt_med,   '| MAD =', percent.mt_mad,   '\n')

nCount_mad_thresh   <- build_mad_table(seurat$nCount_RNA,   nCount_RNA_med,   nCount_RNA_mad)
nFeature_mad_thresh <- build_mad_table(seurat$nFeature_RNA, nFeature_RNA_med, nFeature_RNA_mad)
percent.mt_mad_thresh <- build_mad_table(seurat$percent.mt, percent.mt_med,   percent.mt_mad)
```

```{r compute-mad-adt, eval=isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat)}
nCount_ADT_med <- median(seurat$nCount_ADT, na.rm = TRUE)
nCount_ADT_mad <- median(abs(seurat$nCount_ADT - nCount_ADT_med), na.rm = TRUE)
nCount_ADT_mad_thresh <- build_mad_table(seurat$nCount_ADT, nCount_ADT_med, nCount_ADT_mad)
cat('nCount_ADT median =', nCount_ADT_med, '| MAD =', nCount_ADT_mad, '\n')
```

## UMI count filter table
Each row shows the UMI threshold at a given MAD multiple with the fraction and count of cells that fall below (would be removed as low-quality) or above it.
```{r mad-table-umi}
kable(nCount_mad_thresh, digits = 3,
      caption = 'MAD-based filtering thresholds – UMI count',
      align = c('l', 'c', 'c', 'c', 'c', 'c'))
```

```{r mad-table-adt, eval=isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat)}
kable(nCount_ADT_mad_thresh, digits = 3,
  caption = 'MAD-based filtering thresholds – ADT UMI count',
  align = c('l', 'c', 'c', 'c', 'c', 'c'))
```

## Genes detected filter table
Each row shows the gene-count threshold at a given MAD multiple with the fraction and count of cells falling below or above it.
```{r mad-table-genes}
kable(nFeature_mad_thresh, digits = 3,
      caption = 'MAD-based filtering thresholds – genes detected',
      align = c('l', 'c', 'c', 'c', 'c', 'c'))
```

## Mitochondrial alignment fraction filter table
Each row shows the mitochondrial fraction threshold at a given MAD multiple with the fraction and count of cells falling below or above it.
```{r mad-table-mito}
kable(percent.mt_mad_thresh, digits = 3,
      caption = 'MAD-based filtering thresholds – mitochondrial fraction',
      align = c('l', 'c', 'c', 'c', 'c', 'c'))
```

# QC plots (pre-filter)
Solid red line = median; dashed red lines = ±1–5 MADs from the median.

```{r helper-violin-refs, include=FALSE}
# Helper: build a data frame of reference-line y-positions for violin plots.
# Returns one row per MAD multiple with the y value and linetype
# ("solid" for the median, "dashed" for all other multiples).
# Set log_transform = TRUE for log10-scale metrics (nCount, nFeature).
make_violin_refs <- function(med, mad, log_transform = TRUE) {
  mads      <- -5:5
  vals      <- med + mads * mad
  linetypes <- ifelse(mads == 0, 'solid', 'dashed')
  if (log_transform) {
    keep      <- vals > 0
    vals      <- log10(vals[keep])
    linetypes <- linetypes[keep]
  }
  data.frame(y = vals, linetype = linetypes, stringsAsFactors = FALSE)
}

umi_refs  <- make_violin_refs(nCount_RNA_med,   nCount_RNA_mad,   log_transform = TRUE)
gene_refs <- make_violin_refs(nFeature_RNA_med, nFeature_RNA_mad, log_transform = TRUE)
mt_refs   <- make_violin_refs(percent.mt_med,   percent.mt_mad,   log_transform = FALSE)
if (isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat)) {
  adt_refs <- make_violin_refs(nCount_ADT_med, nCount_ADT_mad, log_transform = TRUE)
}
```

```{r qc-violins-prefilter, fig.height=5, fig.width=12}
# Violin plots for UMI count, gene count, and mitochondrial fraction.
# geom_segment with inherit.aes=FALSE and explicit yend draws correct horizontal tick marks.
cols <- viridis(3)

umi_vio <- ggplot(seurat@meta.data,
                  aes(y = log10(nCount_RNA), x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[1]) + theme(legend.position = 'none') +
  geom_segment(data = umi_refs,
               aes(y = y, yend = y, x = 0.75, xend = 1.25, linetype = linetype),
               color = 'red', linewidth = 0.5, inherit.aes = FALSE) +
  scale_linetype_identity() +
  labs(y = 'log10(UMI count)')

gene_vio <- ggplot(seurat@meta.data,
                   aes(y = log10(nFeature_RNA), x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[2]) + theme(legend.position = 'none') +
  geom_segment(data = gene_refs,
               aes(y = y, yend = y, x = 0.75, xend = 1.25, linetype = linetype),
               color = 'red', linewidth = 0.5, inherit.aes = FALSE) +
  scale_linetype_identity() +
  labs(y = 'log10(genes detected)')

mt_vio <- ggplot(seurat@meta.data,
                 aes(y = percent.mt, x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[3]) + theme(legend.position = 'none') +
  geom_segment(data = mt_refs,
               aes(y = y, yend = y, x = 0.75, xend = 1.25, linetype = linetype),
               color = 'red', linewidth = 0.5, inherit.aes = FALSE) +
  scale_linetype_identity() +
  labs(y = '% mitochondrial')

umi_vio + gene_vio + mt_vio
```

```{r qc-histograms-prefilter, fig.height=4, fig.width=16}
# Histograms of QC metric distributions; positive MAD thresholds shown as dashed lines
umi_hist <- ggplot(seurat@meta.data, aes(x = log10(nCount_RNA), fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[1]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nCount_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nCount_mad_thresh$filter_threshold[nCount_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.5)

gene_hist <- ggplot(seurat@meta.data, aes(x = log10(nFeature_RNA), fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[2]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nFeature_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nFeature_mad_thresh$filter_threshold[nFeature_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.5)

mt_hist <- ggplot(seurat@meta.data, aes(x = percent.mt, fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[3]) + theme(legend.position = 'none') +
  geom_vline(xintercept = percent.mt_med, color = 'red') +
  geom_vline(xintercept = percent.mt_mad_thresh$filter_threshold[percent.mt_mad_thresh$filter_threshold > 0],
             color = 'red', linetype = 'dashed', linewidth = 0.5)

umi_hist + gene_hist + mt_hist
```

```{r qc-scatter-prefilter, fig.height=4, fig.width=16}
# Scatter plots of metric pairs to detect co-varying outliers (e.g., high UMI + high MT)
count_gene <- ggplot(seurat@meta.data,
                     aes(x = log10(nCount_RNA), y = log10(nFeature_RNA), color = orig.ident)) +
  geom_point() + theme_classic() +
  scale_color_manual(values = cols[1]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nCount_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nCount_mad_thresh$filter_threshold[nCount_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25) +
  geom_hline(yintercept = log10(nFeature_RNA_med), color = 'red') +
  geom_hline(yintercept = log10(nFeature_mad_thresh$filter_threshold[nFeature_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25)

count_mito <- ggplot(seurat@meta.data,
                     aes(x = log10(nCount_RNA), y = percent.mt, color = orig.ident)) +
  geom_point() + theme_classic() +
  scale_color_manual(values = cols[2]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nCount_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nCount_mad_thresh$filter_threshold[nCount_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25) +
  geom_hline(yintercept = percent.mt_med, color = 'red') +
  geom_hline(yintercept = percent.mt_mad_thresh$filter_threshold[percent.mt_mad_thresh$filter_threshold > 0],
             color = 'red', linetype = 'dashed', linewidth = 0.25)

gene_mito <- ggplot(seurat@meta.data,
                    aes(x = log10(nFeature_RNA), y = percent.mt, color = orig.ident)) +
  geom_point() + theme_classic() +
  scale_color_manual(values = cols[3]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nFeature_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nFeature_mad_thresh$filter_threshold[nFeature_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25) +
  geom_hline(yintercept = percent.mt_med, color = 'red') +
  geom_hline(yintercept = percent.mt_mad_thresh$filter_threshold[percent.mt_mad_thresh$filter_threshold > 0],
             color = 'red', linetype = 'dashed', linewidth = 0.25)

count_gene + count_mito + gene_mito
```

```{r qc-adt-violin-prefilter, eval=isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat), fig.height=5, fig.width=6}
ggplot(seurat@meta.data,
       aes(y = log10(pmax(nCount_ADT, 1)), x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = viridis(1)) + theme(legend.position = 'none') +
  geom_segment(data = adt_refs,
               aes(y = y, yend = y, x = 0.75, xend = 1.25, linetype = linetype),
               color = 'red', linewidth = 0.5, inherit.aes = FALSE) +
  scale_linetype_identity() +
  labs(y = 'log10(ADT UMI count)')
```

```{r qc-adt-histogram-prefilter, eval=isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat), fig.height=4, fig.width=8}
ggplot(seurat@meta.data, aes(x = log10(pmax(nCount_ADT, 1)))) +
  geom_histogram() + theme_classic() +
  geom_vline(xintercept = log10(pmax(nCount_ADT_med, 1)), color = 'red') +
  geom_vline(xintercept = log10(pmax(nCount_ADT_mad_thresh$filter_threshold[
    nCount_ADT_mad_thresh$filter_threshold > 0], 1)),
    color = 'red', linetype = 'dashed', linewidth = 0.5) +
  labs(x = 'log10(ADT UMI count)')
```

```{r qc-adt-scatter-prefilter, eval=isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat), fig.height=5, fig.width=7}
ggplot(seurat@meta.data,
       aes(x = log10(pmax(nCount_RNA, 1)), y = log10(pmax(nCount_ADT, 1)))) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_vline(xintercept = log10(pmax(nCount_RNA_med, 1)), color = 'red') +
  geom_hline(yintercept = log10(pmax(nCount_ADT_med, 1)), color = 'red') +
  theme_classic() +
  labs(x = 'log10(RNA UMI count)', y = 'log10(ADT UMI count)')
```

## Ribosomal and complexity metrics
Adds ribosomal percentage and visualizes its relationship with mitochondrial
percentage, UMI count, and genes detected for this sample.

```{r qc-ribosomal-prefilter, fig.height=5, fig.width=10}
ribosomal_features <- grep('^(RPL|RPS|Rpl|Rps)', rownames(seurat), value = TRUE)
if (length(ribosomal_features) > 0L) {
  seurat$percent.Ribosomal <- PercentageFeatureSet(
    seurat,
    features = ribosomal_features,
    assay = 'RNA'
  )
} else {
  seurat$percent.Ribosomal <- 0
  cat('No ribosomal genes detected; setting percent.Ribosomal to 0.\n')
}

VlnPlot(
  seurat,
  features = c('percent.Ribosomal', 'percent.mt'),
  group.by = 'orig.ident',
  pt.size = 0,
  log = TRUE,
  ncol = 2
) + NoLegend()
```

```{r qc-ribosomal-scatter-prefilter, fig.height=8, fig.width=12}
qc.metrics <- seurat@meta.data

ribo_scatter <- ggplot(
  qc.metrics,
  aes(x = nCount_RNA, y = nFeature_RNA, colour = percent.Ribosomal)
) +
  geom_point(size = 0.5) +
  scale_color_viridis_c() +
  scale_x_log10() +
  scale_y_log10() +
  theme_classic() +
  labs(
    title = 'UMI count versus genes detected',
    colour = '% ribosomal'
  )

mt_scatter <- ggplot(
  qc.metrics,
  aes(x = nCount_RNA, y = nFeature_RNA, colour = percent.mt)
) +
  geom_point(size = 0.5) +
  scale_color_viridis_c() +
  scale_x_log10() +
  scale_y_log10() +
  theme_classic() +
  labs(
    title = 'UMI count versus genes detected',
    colour = '% mitochondrial'
  )

print(ribo_scatter)
print(mt_scatter)
```

```{r qc-complexity-prefilter, fig.height=5, fig.width=10}
qc.metrics$complexity <- with(
  qc.metrics,
  log10(nFeature_RNA) / log10(nCount_RNA)
)

complexity.lm <- lm(
  log10(nFeature_RNA) ~ log10(nCount_RNA),
  data = qc.metrics
)

qc.metrics$complexity_diff <- with(
  qc.metrics,
  log10(nFeature_RNA) - predict(complexity.lm, newdata = qc.metrics)
)

complexity_plot <- ggplot(
  qc.metrics,
  aes(x = log10(nCount_RNA), y = log10(nFeature_RNA), colour = complexity_diff)
) +
  geom_point(size = 0.5) +
  geom_abline(
    slope = coef(complexity.lm)[2],
    intercept = coef(complexity.lm)[1]
  ) +
  scale_colour_gradient2(low = 'blue2', mid = 'grey', high = 'red2') +
  theme_classic() +
  labs(
    title = 'Cell complexity',
    x = 'log10(UMI count)',
    y = 'log10(genes detected)',
    colour = 'Complexity residual'
  )

complexity_qc_plot <- ggplot(
  qc.metrics,
  aes(x = complexity_diff, y = percent.Ribosomal, colour = percent.mt)
) +
  geom_point(size = 1) +
  scale_color_viridis_c() +
  theme_classic() +
  labs(
    title = 'Complexity residual versus QC percentages',
    x = 'Complexity residual',
    y = '% ribosomal',
    colour = '% mitochondrial'
  )

print(complexity_plot)
print(complexity_qc_plot)
```

# MALAT1
MALAT1 is a highly-expressed nuclear lncRNA. After normalisation, cells with very low MALAT1 are likely lysed or damaged. The feature name (`MALAT1` / `Malat1`) is selected automatically based on the detected species.

```{r malat1-scatter, fig.height=5, fig.width=7}
# MALAT1 is a nuclear lncRNA; low normalised MALAT1 indicates poor-quality or
# lysed cells. Normalise the data before plotting.
seurat <- NormalizeData(seurat, assay = 'RNA')
if (is.na(malat1_feature)) {
  cat('Skipping MALAT1 scatter: MALAT1/ Malat1 not found in feature set.\n')
} else {
  scatter_df <- FetchData(seurat, vars = c('nCount_RNA', malat1_feature), layer = 'data')
  colnames(scatter_df) <- c('nCount_RNA', 'MALAT1')
  ggplot(scatter_df, aes(x = nCount_RNA, y = MALAT1, color = nCount_RNA)) +
    geom_point(size = 0.5, alpha = 0.5) +
    scale_color_viridis_c(name = 'nCount_RNA') +
    theme_classic() +
    labs(x = 'UMI count (nCount_RNA)', y = malat1_feature,
         title = 'MALAT1 vs UMI count — colored by nCount_RNA')
}
```

# Red blood cells (pre-filter)
Haemoglobin gene expression is used as a proxy for residual red blood cell (RBC) contamination. The species-specific gene list from `human_hemoglobin_genes.csv` / `mouse_hemoglobin_genes.csv` is intersected with detected features.

```{r rbc-prefilter, fig.width=4, fig.height=3}
# Haemoglobin genes flag RBC contamination; species-specific feature set.
if (length(rbc_features) > 0) {
  seurat[["percent.rbc"]] <- PercentageFeatureSet(seurat, features = rbc_features)
} else {
  seurat[["percent.rbc"]] <- 0
  cat('No RBC genes detected for species', species, '- setting percent.rbc to 0.\n')
}

ggplot(seurat@meta.data, aes(y = percent.rbc, x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[1]) + theme(legend.position = 'none')

ggplot(seurat@meta.data, aes(x = percent.rbc, fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[3]) + theme(legend.position = 'none')
```

```{r summary-rbc}
print(describe(seurat$percent.rbc))
```

# Filter
Applies per-metric RNA and optional ADT thresholds to remove low-quality cells. Uses hard thresholds from `params` if provided (non-`NA`); otherwise falls back to MAD-based defaults (−4 MADs for count/gene lower bounds, +5 MADs for mitochondrial upper bound). MALAT1 filtering is applied on the normalised `data` layer and skipped if the gene is absent.

```{r apply-filter}
# ── Resolve filter thresholds ──────────────────────────────────────────────────
# For each metric, use the hard threshold from params if supplied (non-NA);
# otherwise fall back to the MAD-based default (±4 MADs for counts/genes,
# +5 MADs for mitochondrial fraction).
resolve_thresh <- function(param_val, mad_val) {
  if (!is.na(param_val)) param_val else mad_val
}

thresh_min_nCount   <- resolve_thresh(params$min_nCount_RNA,
                         nCount_mad_thresh$filter_threshold[nCount_mad_thresh$mad == -4])
thresh_max_nCount   <- resolve_thresh(params$max_nCount_RNA,   Inf)
thresh_min_nFeature <- resolve_thresh(params$min_nFeature_RNA,
                         nFeature_mad_thresh$filter_threshold[nFeature_mad_thresh$mad == -4])
thresh_max_nFeature <- resolve_thresh(params$max_nFeature_RNA, Inf)
thresh_max_mt       <- resolve_thresh(params$max_percent_mt,
                         percent.mt_mad_thresh$filter_threshold[percent.mt_mad_thresh$mad == 5])
thresh_min_malat1   <- resolve_thresh(params$min_malat1, 1)

# Guard against NA mitochondrial thresholds (seen in some CellBender outputs)
# so QC filtering does not drop all cells due to logical NA comparisons.
if (is.na(thresh_max_mt)) {
  thresh_max_mt <- Inf
  cat('percent.mt threshold was NA; using Inf (no mitochondrial upper-bound filter).\n')
}

cat('Filter thresholds applied:\n')
cat('  nCount_RNA   :', thresh_min_nCount,   '–', thresh_max_nCount,   '\n')
cat('  nFeature_RNA :', thresh_min_nFeature, '–', thresh_max_nFeature, '\n')
cat('  percent.mt   : <=', thresh_max_mt,  '\n')
cat('  MALAT1 (norm): >', thresh_min_malat1, '\n')

has_adt_qc <- isTRUE(params$use_adt) && 'ADT' %in% Assays(seurat)
if (has_adt_qc) {
  adt_count_thresholds <- resolve_adt_count_thresholds(
    seurat$nCount_ADT,
    min_override = params$min_nCount_ADT,
    max_override = params$max_nCount_ADT
  )
  pass_nCount_ADT <- adt_count_keep_mask(
    seurat$nCount_ADT,
    adt_count_thresholds$min,
    adt_count_thresholds$max
  )
  cat('  nCount_ADT   :', adt_count_thresholds$min, '–', adt_count_thresholds$max, '\n')
} else {
  pass_nCount_ADT <- rep(TRUE, ncol(seurat))
  cat('  nCount_ADT   : not applicable\n')
}

# Log pass counts per criterion to make over-filtering easier to diagnose.
pass_nCount_min <- seurat$nCount_RNA >  thresh_min_nCount
pass_nCount_max <- seurat$nCount_RNA <= thresh_max_nCount
pass_nFeat_min  <- seurat$nFeature_RNA >  thresh_min_nFeature
pass_nFeat_max  <- seurat$nFeature_RNA <= thresh_max_nFeature
pass_mt_max     <- is.na(seurat$percent.mt) | seurat$percent.mt <= thresh_max_mt

cat('Cells passing each criterion before combined filter:\n')
cat('  nCount_RNA   > min :', sum(pass_nCount_min), '/', length(pass_nCount_min), '\n')
cat('  nCount_RNA   <= max:', sum(pass_nCount_max), '/', length(pass_nCount_max), '\n')
cat('  nFeature_RNA > min :', sum(pass_nFeat_min),  '/', length(pass_nFeat_min),  '\n')
cat('  nFeature_RNA <= max:', sum(pass_nFeat_max),  '/', length(pass_nFeat_max),  '\n')
cat('  percent.mt   <= max:', sum(pass_mt_max),     '/', length(pass_mt_max),     '\n')
cat('  nCount_ADT   in range:', sum(pass_nCount_ADT), '/', length(pass_nCount_ADT), '\n')

# ── Apply filter ───────────────────────────────────────────────────────────────
keep_cells <- colnames(seurat)[
  seurat$nCount_RNA > 0 & seurat$nFeature_RNA > 0 &
    pass_nCount_min & pass_nCount_max & pass_nFeat_min & pass_nFeat_max &
    pass_mt_max & pass_nCount_ADT
]
seurat <- subset(seurat, cells = keep_cells)
if (!is.na(malat1_feature)) {
  malat_vals <- Seurat::FetchData(seurat, vars = malat1_feature, layer = 'data')[, 1]
  keep_cells <- names(malat_vals)[malat_vals > thresh_min_malat1]
  seurat <- subset(seurat, cells = keep_cells)
} else {
  cat('Skipping MALAT1-based filtering: MALAT1/ Malat1 not found.\n')
}
```

# Cell count post-filter
Prints the Seurat object summary after all filters have been applied to confirm the retained cell and feature counts.

```{r post-filter-object-summary}
seurat
```

# QC plots (post-filter)
Solid red line = pre-filter median; dashed red lines = pre-filter MAD thresholds (shown as reference).

```{r qc-violins-postfilter, fig.height=5, fig.width=12}
# Same reference data frames (umi_refs, gene_refs, mt_refs) as pre-filter for direct comparison
umi_vio <- ggplot(seurat@meta.data,
                  aes(y = log10(nCount_RNA), x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[1]) + theme(legend.position = 'none') +
  geom_segment(data = umi_refs,
               aes(y = y, yend = y, x = 0.75, xend = 1.25, linetype = linetype),
               color = 'red', linewidth = 0.5, inherit.aes = FALSE) +
  scale_linetype_identity() +
  labs(y = 'log10(UMI count)')

gene_vio <- ggplot(seurat@meta.data,
                   aes(y = log10(nFeature_RNA), x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[2]) + theme(legend.position = 'none') +
  geom_segment(data = gene_refs,
               aes(y = y, yend = y, x = 0.75, xend = 1.25, linetype = linetype),
               color = 'red', linewidth = 0.5, inherit.aes = FALSE) +
  scale_linetype_identity() +
  labs(y = 'log10(genes detected)')

mt_vio <- ggplot(seurat@meta.data,
                 aes(y = percent.mt, x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[3]) + theme(legend.position = 'none') +
  geom_segment(data = mt_refs,
               aes(y = y, yend = y, x = 0.75, xend = 1.25, linetype = linetype),
               color = 'red', linewidth = 0.5, inherit.aes = FALSE) +
  scale_linetype_identity() +
  labs(y = '% mitochondrial')

umi_vio + gene_vio + mt_vio
```

```{r qc-histograms-postfilter, fig.height=4, fig.width=16}
# Histograms after filtering — same MAD reference lines as pre-filter
umi_hist <- ggplot(seurat@meta.data, aes(x = log10(nCount_RNA), fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[1]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nCount_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nCount_mad_thresh$filter_threshold[nCount_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.5)

gene_hist <- ggplot(seurat@meta.data, aes(x = log10(nFeature_RNA), fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[2]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nFeature_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nFeature_mad_thresh$filter_threshold[nFeature_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.5)

mt_hist <- ggplot(seurat@meta.data, aes(x = percent.mt, fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[3]) + theme(legend.position = 'none') +
  geom_vline(xintercept = percent.mt_med, color = 'red') +
  geom_vline(xintercept = percent.mt_mad_thresh$filter_threshold[percent.mt_mad_thresh$filter_threshold > 0],
             color = 'red', linetype = 'dashed', linewidth = 0.5)

umi_hist + gene_hist + mt_hist
```

```{r qc-scatter-postfilter, fig.height=4, fig.width=16}
# Scatter plots after filtering
count_gene <- ggplot(seurat@meta.data,
                     aes(x = log10(nCount_RNA), y = log10(nFeature_RNA), color = orig.ident)) +
  geom_point() + theme_classic() +
  scale_color_manual(values = cols[1]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nCount_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nCount_mad_thresh$filter_threshold[nCount_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25) +
  geom_hline(yintercept = log10(nFeature_RNA_med), color = 'red') +
  geom_hline(yintercept = log10(nFeature_mad_thresh$filter_threshold[nFeature_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25)

count_mito <- ggplot(seurat@meta.data,
                     aes(x = log10(nCount_RNA), y = percent.mt, color = orig.ident)) +
  geom_point() + theme_classic() +
  scale_color_manual(values = cols[2]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nCount_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nCount_mad_thresh$filter_threshold[nCount_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25) +
  geom_hline(yintercept = percent.mt_med, color = 'red') +
  geom_hline(yintercept = percent.mt_mad_thresh$filter_threshold[percent.mt_mad_thresh$filter_threshold > 0],
             color = 'red', linetype = 'dashed', linewidth = 0.25)

gene_mito <- ggplot(seurat@meta.data,
                    aes(x = log10(nFeature_RNA), y = percent.mt, color = orig.ident)) +
  geom_point() + theme_classic() +
  scale_color_manual(values = cols[3]) + theme(legend.position = 'none') +
  geom_vline(xintercept = log10(nFeature_RNA_med), color = 'red') +
  geom_vline(xintercept = log10(nFeature_mad_thresh$filter_threshold[nFeature_mad_thresh$filter_threshold > 0]),
             color = 'red', linetype = 'dashed', linewidth = 0.25) +
  geom_hline(yintercept = percent.mt_med, color = 'red') +
  geom_hline(yintercept = percent.mt_mad_thresh$filter_threshold[percent.mt_mad_thresh$filter_threshold > 0],
             color = 'red', linetype = 'dashed', linewidth = 0.25)

count_gene + count_mito + gene_mito
```

# Dimensionality reduction & clustering
Runs the standard Seurat RNA workflow on the filtered object: highly variable feature selection → scaling → PCA → SNN graph → Leiden clustering → UMAP embedding. The elbow plot is printed to help assess how many PCs to retain.

```{r dim-reduction-clustering}
# Standard Seurat preprocessing: variable features → scale → PCA → graph → cluster → UMAP.
# NormalizeData was already called in the MALAT1 section above; not repeated here.
seurat <- FindVariableFeatures(seurat, assay = 'RNA')
seurat <- ScaleData(seurat, assay = 'RNA')
seurat <- RunPCA(seurat, assay = 'RNA', reduction.name = 'pca_rna')
ElbowPlot(seurat, reduction = 'pca_rna')
seurat <- FindNeighbors(seurat, assay = 'RNA', reduction = 'pca_rna', dims = 1:30)
seurat <- FindClusters(seurat, algorithm = 4, cluster.name = 'seurat_clusters_rna')
seurat <- RunUMAP(seurat, dims = 1:30, reduction = 'pca_rna', reduction.name = 'umap_rna')
```

# Most variable genes (log1p)
Mean–variance plot of log-normalised expression with the top 30 most variable genes labelled. Assess whether variable features represent biology or technical noise.

```{r variable-features-plot, fig.height=8, fig.width=10}
# Label the top 30 most variable genes on the mean-variance plot
top30 <- head(VariableFeatures(seurat), 30)
plot1 <- VariableFeaturePlot(seurat)
plot2 <- LabelPoints(plot = plot1, points = top30, repel = TRUE)
plot2
```

# Mean–variance relationship
In scRNA-seq count data, variance typically scales with the mean (higher-expressed genes are more variable). Three plots are shown:

1. **Log mean vs log variance**: Points falling above the diagonal red line (slope = 1, the Poisson expectation) are overdispersed — they vary more than expected by chance alone. Overdispersed genes carry the most biological signal and drive variable feature selection.
2. **Detection rate vs log mean**: The red curve shows the Poisson-predicted detection rate at each mean expression level. Genes that fall below the curve are detected less often than expected, which can indicate bursty or bimodal expression. Genes above the curve are more consistently detected.
3. **UMI count vs genes detected per cell**: Cells should follow a roughly linear relationship between total UMI count and number of genes detected. Outliers — cells with many UMIs but few genes, or many genes but few UMIs — may represent doublets, empty droplets, or poor-quality cells that survived earlier filters.

These plots motivate the use of SCTransform downstream, which directly models and removes the mean–variance trend to prevent highly expressed housekeeping genes from dominating dimensionality reduction.

```{r mean-variance, fig.height=4, fig.width=8}
# Compute per-gene and per-cell summary statistics from the raw count matrix
gene_attr <- data.frame(
  mean           = rowMeans(seurat@assays$RNA$counts),
  detection_rate = rowMeans(seurat@assays$RNA$counts > 0),
  var            = apply(seurat@assays$RNA$counts, 1, var)
)
gene_attr$log_mean <- log10(gene_attr$mean)
gene_attr$log_var  <- log10(gene_attr$var)
rownames(gene_attr) <- rownames(seurat)

cell_attr <- data.frame(
  n_umi  = colSums(seurat@assays$RNA$counts),
  n_gene = colSums(seurat@assays$RNA$counts > 0)
)
rownames(cell_attr) <- colnames(seurat@assays$RNA)

# Mean-variance: points above the diagonal (red line) show overdispersion beyond Poisson
ggplot(gene_attr, aes(log_mean, log_var)) +
  geom_point(alpha = 0.3, shape = 16) +
  geom_density_2d(linewidth = 0.3) +
  geom_abline(intercept = 0, slope = 1, color = 'red')

# Detection rate vs mean: deviation from the Poisson prediction (red) suggests overdispersion
x <- seq(from = -3, to = 2, length.out = 1000)
poisson_model <- data.frame(
  log_mean       = x,
  detection_rate = 1 - dpois(0, lambda = 10^x)
)
ggplot(gene_attr, aes(log_mean, detection_rate)) +
  geom_point(alpha = 0.3, shape = 16) +
  geom_line(data = poisson_model, color = 'red') +
  theme_gray(base_size = 8)

# UMI vs gene count per cell
ggplot(cell_attr, aes(n_umi, n_gene)) +
  geom_point(alpha = 0.3, shape = 16) +
  geom_density_2d(linewidth = 0.3)
```

## SCTransform per-gene and regularised model coefficients

```{r sctransform-model, fig.height=4, fig.width=16}
# Fit SCTransform model to evaluate variance-stabilisation quality.
# future.globals.maxSize set to 8 GB to accommodate large count matrices.
# Guard against zero-UMI cells, which produce -Inf log_umi and break sctransform::vst.
zero_umi_cells <- colnames(seurat)[seurat$nCount_RNA <= 0 | is.na(seurat$nCount_RNA)]
if (length(zero_umi_cells) > 0) {
  cat('Removing', length(zero_umi_cells), 'cells with nCount_RNA <= 0 before SCTransform model fitting.\n')
  seurat <- subset(seurat, cells = setdiff(colnames(seurat), zero_umi_cells))
}
if (ncol(seurat) == 0) {
  stop('No cells remain after removing zero-UMI cells prior to SCTransform.', call. = FALSE)
}

set.seed(44)
options(future.globals.maxSize = 64000 * 1024^2) 
vst_out <- sctransform::vst(
  seurat@assays$RNA$counts,
  latent_var       = c('log_umi'),
  return_gene_attr = TRUE,
  return_cell_attr = TRUE,
  verbosity        = 1,
  vst.flavor       = "v2"
)
sctransform::plot_model_pars(vst_out, show_theta = TRUE) + theme_bw()
```

# SCTransform normalization workflow

```{r sctransform-normalize}
# Normalize with SCTransform and rerun dimensionality reduction
# on the SCT assay. 
seurat <- SCTransform(seurat, assay = 'RNA', new.assay.name = 'SCT', verbose = FALSE, vst.flavor = "v2")

seurat <- RunPCA(seurat, assay = 'SCT', reduction.name = 'pca_sct')
seurat <- FindNeighbors(seurat, reduction = 'pca_sct', dims = 1:30)
seurat <- FindClusters(seurat, algorithm = 4, cluster.name = 'seurat_clusters_sct')
seurat <- RunUMAP(seurat, reduction = 'pca_sct', dims = 1:30, reduction.name = 'umap_sct')

# Keep a default cluster column for compatibility with downstream tooling.
seurat$seurat_clusters <- seurat$seurat_clusters_sct
Idents(seurat) <- 'seurat_clusters_sct'
```

# UMAP plots by normalisation method
Two-dimensional UMAP embeddings for RNA (log-normalised) and SCTransform workflows.

```{r umap-plot}
DimPlot(seurat, reduction = 'umap_rna', group.by = 'seurat_clusters_rna', label = TRUE, repel = TRUE) +
  ggtitle('UMAP (LogNormalize method)')

DimPlot(seurat, reduction = 'umap_sct', group.by = 'seurat_clusters_sct', label = TRUE, repel = TRUE) +
  ggtitle('UMAP (SCTransform normalisation)')
```

# Graph modularity by normalisation method
Computes weighted graph modularity for both RNA (log-normalized) and SCTransform SNN graphs using their respective cluster labels, then visualises pairwise modularity ratios with a heatmap and a cluster network where edge thickness scales with modularity ratio.

The following description was pulled from the OCSA with bioconductor notebook:
For graph-based clustering, the modularity is a natural metric for evaluating the separation between communities. This is defined as the (scaled) difference between the observed total weight of edges between nodes in the same cluster and the expected total weight if edge weights were randomly distributed across all pairs of nodes. Larger modularity values indicate that there most edges occur within clusters, suggesting that the clusters are sufficiently well separated to avoid edges forming between neighboring cells in different clusters.

The standard approach is to report a single modularity value for a clustering on a given graph. This is useful for comparing different clusterings on the same graph - and indeed, some community detection algorithms are designed with the aim of maximizing the modularity - but it is less helpful for interpreting a given clustering. Rather, we use the pairwiseModularity() function from bluster with as.ratio=TRUE, which returns the ratio of the observed to expected sum of weights between each pair of clusters. We use the ratio instead of the difference as the former is less dependent on the number of cells in each cluster.

```{r graph-modularity, fig.height=7, fig.width=9}
if (!requireNamespace('bluster', quietly = TRUE)) {
  stop('The bluster package is required for pairwiseModularity. Please install it before rendering.', call. = FALSE)
}

plot_method_modularity <- function(obj, graph_name, cluster_col, method_label) {
  if (!graph_name %in% names(obj@graphs)) {
    stop(graph_name, ' graph not found in seurat@graphs; cannot compute ', method_label, ' graph modularity.', call. = FALSE)
  }

  graph_mat <- obj@graphs[[graph_name]]
  graph_mat <- Matrix::triu(graph_mat, k = 1)
  edge_triplets <- summary(graph_mat)

  edge_df <- data.frame(
    from = colnames(graph_mat)[edge_triplets$i],
    to = colnames(graph_mat)[edge_triplets$j],
    weight = edge_triplets$x,
    stringsAsFactors = FALSE
  )

  g <- igraph::graph_from_data_frame(
    edge_df,
    directed = FALSE,
    vertices = colnames(graph_mat)
  )

  cluster_labels <- factor(as.character(obj[[cluster_col]][colnames(graph_mat), 1]))
  if (any(is.na(cluster_labels))) {
    stop('Missing ', method_label, ' cluster labels for one or more graph vertices.', call. = FALSE)
  }

  global_modularity <- igraph::modularity(
    g,
    membership = as.integer(cluster_labels),
    weights = igraph::E(g)$weight
  )
  cat('Global weighted modularity (', graph_name, ', ', cluster_col, '): ', signif(global_modularity, 5), '\n', sep = '')

  # Pairwise modularity ratios (>1 enriched connectivity, <1 depleted).
  pair_mod_ratio <- bluster::pairwiseModularity(g, cluster_labels, as.ratio = TRUE)
  pair_mod_ratio <- as.matrix(pair_mod_ratio)

  heat_df <- as.data.frame(as.table(pair_mod_ratio), stringsAsFactors = FALSE)
  colnames(heat_df) <- c('cluster1', 'cluster2', 'modularity_ratio')
  heat_df$cluster1 <- factor(heat_df$cluster1, levels = sort(unique(heat_df$cluster1)))
  heat_df$cluster2 <- factor(heat_df$cluster2, levels = sort(unique(heat_df$cluster2)))
  heat_df$modularity_ratio_plot <- ifelse(
    is.finite(heat_df$modularity_ratio) & heat_df$modularity_ratio > 0,
    heat_df$modularity_ratio,
    NA_real_
  )

  print(
    ggplot(heat_df, aes(x = cluster1, y = cluster2, fill = modularity_ratio_plot)) +
      geom_tile() +
      scale_fill_viridis_c(option = 'magma', trans = 'log10', na.value = 'grey90') +
      theme_classic() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      xlab(paste(method_label, 'cluster')) +
      ylab(paste(method_label, 'cluster')) +
      ggtitle(paste('Pairwise modularity ratio heatmap (', method_label, ')', sep = ''))
  )

  # Cluster-level network: edge thickness scales with pairwise modularity ratio.
  upper_idx <- which(upper.tri(pair_mod_ratio), arr.ind = TRUE)
  net_edges <- data.frame(
    from = rownames(pair_mod_ratio)[upper_idx[, 1]],
    to = colnames(pair_mod_ratio)[upper_idx[, 2]],
    modularity_ratio = pair_mod_ratio[upper_idx],
    stringsAsFactors = FALSE
  )
  net_edges <- net_edges[is.finite(net_edges$modularity_ratio) & net_edges$modularity_ratio > 0, , drop = FALSE]

  if (nrow(net_edges) > 0) {
    cluster_nodes <- sort(unique(c(net_edges$from, net_edges$to)))
    g_clusters <- igraph::graph_from_data_frame(
      net_edges,
      directed = FALSE,
      vertices = data.frame(name = cluster_nodes, stringsAsFactors = FALSE)
    )

    lay <- igraph::layout_with_fr(g_clusters, weights = igraph::E(g_clusters)$modularity_ratio)
    node_df <- data.frame(
      cluster = igraph::V(g_clusters)$name,
      x = lay[, 1],
      y = lay[, 2],
      stringsAsFactors = FALSE
    )

    e_tbl <- igraph::as_data_frame(g_clusters, what = 'edges')
    e_tbl$x <- node_df$x[match(e_tbl$from, node_df$cluster)]
    e_tbl$y <- node_df$y[match(e_tbl$from, node_df$cluster)]
    e_tbl$xend <- node_df$x[match(e_tbl$to, node_df$cluster)]
    e_tbl$yend <- node_df$y[match(e_tbl$to, node_df$cluster)]

    print(
      ggplot() +
        geom_segment(
          data = e_tbl,
          aes(x = x, y = y, xend = xend, yend = yend,
              size = modularity_ratio, color = modularity_ratio),
          alpha = 0.8
        ) +
        geom_point(data = node_df, aes(x = x, y = y), color = 'black', size = 3) +
        geom_text(data = node_df, aes(x = x, y = y, label = cluster), vjust = -1, size = 3) +
        scale_color_viridis_c(option = 'plasma', name = 'Modularity ratio') +
        scale_size_continuous(range = c(0.3, 4), name = 'Modularity ratio') +
        theme_void() +
        ggtitle(paste('Cluster network weighted by pairwise modularity ratio (', method_label, ')', sep = ''))
    )
  } else {
    cat('No finite positive pairwise modularity ratios available for ', method_label, ' network plot.\n', sep = '')
  }
}

plot_method_modularity(
  seurat,
  graph_name = 'RNA_snn',
  cluster_col = 'seurat_clusters_rna',
  method_label = 'LogNormalize method'
)

plot_method_modularity(
  seurat,
  graph_name = 'SCT_snn',
  cluster_col = 'seurat_clusters_sct',
  method_label = 'SCT'
)
```

# QC metrics across clusters by normalisation method
Projects log-scale UMI count and gene count onto each UMAP and visualises QC metric distributions by method-specific cluster labels.

```{r qc-by-cluster, fig.height=7, fig.width=10}
# Log-scale QC metrics projected onto RNA and SCT UMAPs and split by
# method-specific cluster identities.
seurat$nCount_RNA_log   <- log10(seurat$nCount_RNA)
seurat$nFeature_RNA_log <- log10(seurat$nFeature_RNA)

safe_vlnplot <- function(obj, group_col, title_text) {
  meta <- obj@meta.data
  if (!group_col %in% colnames(meta)) {
    stop('Grouping column not found in meta.data: ', group_col, call. = FALSE)
  }

  plot_one <- function(feature, ylab) {
    ggplot(meta, aes(x = .data[[group_col]], y = .data[[feature]], fill = .data[[group_col]])) +
      geom_violin(trim = FALSE) +
      theme_classic() +
      theme(legend.position = 'none', axis.text.x = element_text(angle = 45, hjust = 1), axis.text.y = element_text(size = 8)) +
      labs(x = group_col, y = ylab)
  }

  wrap_plots(
    plot_one('nCount_RNA_log', 'log10(nCount_RNA)'),
    plot_one('nFeature_RNA_log', 'log10(nFeature_RNA)'),
    plot_one('percent.mt', '% mitochondrial'),
    ncol = 1
  ) + plot_annotation(title = title_text)
}

FeaturePlot(seurat, features = 'nCount_RNA_log', reduction = 'umap_rna') +
  ggtitle('RNA UMAP: log10 nCount_RNA')
FeaturePlot(seurat, features = 'nFeature_RNA_log', reduction = 'umap_rna') +
  ggtitle('RNA UMAP: log10 nFeature_RNA')
safe_vlnplot(seurat, 'seurat_clusters_rna', 'RNA clusters: QC metrics')

FeaturePlot(seurat, features = 'nCount_RNA_log', reduction = 'umap_sct') +
  ggtitle('SCT UMAP: log10 nCount_RNA')
FeaturePlot(seurat, features = 'nFeature_RNA_log', reduction = 'umap_sct') +
  ggtitle('SCT UMAP: log10 nFeature_RNA')
safe_vlnplot(seurat, 'seurat_clusters_sct', 'SCT clusters: QC metrics')
```

# Red blood cells (post-clustering)
Checks whether any clusters are enriched for haemoglobin gene expression after filtering. Clusters with high `percent.rbc` likely represent residual RBC contamination and may warrant removal.

```{r rbc-postfilter}
# Check whether any remaining clusters are enriched for RBC contamination
FeaturePlot(seurat, features = 'percent.rbc')
ggplot(seurat@meta.data, aes(y = percent.rbc, x = orig.ident, fill = orig.ident)) +
  geom_violin() + theme_classic() +
  scale_fill_manual(values = cols[1]) + theme(legend.position = 'none')
ggplot(seurat@meta.data, aes(x = percent.rbc, fill = orig.ident)) +
  geom_histogram() + theme_classic() +
  scale_fill_manual(values = cols[3]) + theme(legend.position = 'none')
```

# Cell cycle scoring
Scores each cell for S-phase and G2/M-phase activity using Seurat's built-in marker gene lists (`cc.genes.updated.2019`). For mouse data, human symbols are converted to title-case. Scores and phase assignment are stored in cell metadata as `S.Score`, `G2M.Score`, and `Phase`.

```{r cell-cycle-scoring}
# Score each cell for S and G2M phase activity using Seurat's built-in gene lists.
# Built-in cc.genes / cc.genes.updated.2019 are human symbols; for mouse, convert
# to title case (e.g. TOP2A → Top2a) which matches Ensembl mouse gene symbols.
cc <- cc.genes.updated.2019

if (species == 'human') {
  s_genes   <- cc$s.genes
  g2m_genes <- cc$g2m.genes
} else {
  s_genes   <- tools::toTitleCase(tolower(cc$s.genes))
  g2m_genes <- tools::toTitleCase(tolower(cc$g2m.genes))
}

# Restrict to genes actually present in the assay to avoid warnings
s_genes   <- intersect(s_genes,   rownames(seurat))
g2m_genes <- intersect(g2m_genes, rownames(seurat))

cat('S-phase genes found:  ', length(s_genes),   '\n')
cat('G2M-phase genes found:', length(g2m_genes), '\n')

seurat <- CellCycleScoring(
  seurat,
  s.features   = s_genes,
  g2m.features = g2m_genes,
  assay        = 'RNA'
)
```

```{r cell-cycle-umap, fig.height=7, fig.width=10}
# UMAP coloured by phase and by continuous S / G2M scores
phase_plot <- DimPlot(seurat, group.by = 'Phase', label = FALSE) +
  ggtitle('Cell cycle phase')

s_plot   <- FeaturePlot(seurat, features = 'S.Score',   order = TRUE) +
  scale_colour_viridis_c()
g2m_plot <- FeaturePlot(seurat, features = 'G2M.Score', order = TRUE) +
  scale_colour_viridis_c()

phase_plot
s_plot
g2m_plot
```

```{r cell-cycle-violin, fig.height=7, fig.width=12}
# S and G2M scores per cluster; high scores in a cluster suggest cycling cells
s_vln <- VlnPlot(seurat, features = 'S.Score',   group.by = 'seurat_clusters',
                 assay = 'RNA', pt.size = 0) + NoLegend() + xlab('Cluster')
g2m_vln <- VlnPlot(seurat, features = 'G2M.Score', group.by = 'seurat_clusters_sct',
                   assay = 'RNA', pt.size = 0) + NoLegend() + xlab('Cluster')
s_vln
g2m_vln
```

```{r cell-cycle-phase-table}
# Counts and proportions of each phase per cluster for both methods
phase_prop_rna <- as.data.frame(
  prop.table(table(Cluster = seurat$seurat_clusters_rna, Phase = seurat$Phase), margin = 1)
)
phase_prop_sct <- as.data.frame(
  prop.table(table(Cluster = seurat$seurat_clusters_sct, Phase = seurat$Phase), margin = 1)
)

ggplot(phase_prop_rna, aes(x = Cluster, y = Freq, fill = Phase)) +
  geom_bar(stat = 'identity') +
  theme_classic() +
  ylab('Fraction') +
  ggtitle('Cell cycle phase composition per cluster (RNA normalisation)')

ggplot(phase_prop_sct, aes(x = Cluster, y = Freq, fill = Phase)) +
  geom_bar(stat = 'identity') +
  theme_classic() +
  ylab('Fraction') +
  ggtitle('Cell cycle phase composition per cluster (SCTransform normalisation)')
```

# Top marker heatmap
Identifies per-cluster marker genes using the Wilcoxon rank-sum test (presto). The top 20 markers per cluster (log-FC > 0.25, adjusted p < 0.05) are displayed as a row-scaled heatmap of average normalised expression.

```{r marker-heatmap, fig.height=18, fig.width=8}
# Identify per-cluster markers with Wilcoxon test (presto), then display the
# average normalised expression of the top 20 markers per cluster for RNA/SCT.
build_top_marker_heatmap <- function(
  obj,
  cluster_col,
  assay_name,
  layer_name,
  title_text
) {
  markers <- wilcoxauc(
    obj,
    only.pos = TRUE,
    assay = layer_name,
    seurat_assay = assay_name,
    group_by = cluster_col
  )

  top20 <- markers %>%
    group_by(group) %>%
    dplyr::filter(logFC > 0.25 & padj < 0.05) %>%
    slice_head(n = 20) %>%
    ungroup()

  if (nrow(top20) == 0) {
    message('No marker genes passed filters for: ', title_text)
    return(invisible(NULL))
  }

  avg_exp <- AverageExpression(
    obj,
    features = unique(top20$feature),
    assay = assay_name,
    layer = layer_name,
    group.by = cluster_col
  )

  mat <- avg_exp[[assay_name]]
  if (nrow(mat) == 0 || ncol(mat) == 0) {
    message('Empty average expression matrix for: ', title_text)
    return(invisible(NULL))
  }

  pheatmap::pheatmap(
    mat,
    scale = 'row',
    color = myColor,
    border_color = NULL,
    breaks = myBreaks,
    fontsize = 6,
    main = title_text
  )
}

paletteLength <- 50
myColor  <- colorRampPalette(c('blue', 'white', 'red'))(paletteLength)
myBreaks <- c(
  seq(-3, 0,              length.out = ceiling(paletteLength / 2) + 1),
  seq(3 / paletteLength, 3, length.out = floor(paletteLength / 2))
)

build_top_marker_heatmap(
  seurat,
  cluster_col = 'seurat_clusters_rna',
  assay_name = 'RNA',
  layer_name = 'data',
  title_text = 'Top marker heatmap (LogNormalize method)'
)

build_top_marker_heatmap(
  seurat,
  cluster_col = 'seurat_clusters_sct',
  assay_name = 'SCT',
  layer_name = 'data',
  title_text = 'Top marker heatmap (SCTransform normalisation)'
)
```

```{r adt-section-header, results='asis', echo=FALSE}
if (isTRUE(params$use_adt)) {
  cat(
    '# CITE-seq (ADT) processing\n\n',
    'Normalizes CITE-seq surface protein (ADT) counts with `dsb`, using empty ',
    'droplets from the raw matrix as background, and visualizes normalized ',
    'antibody levels across RNA/SCT clusters.\n\n',
    sep = ''
  )
}
```

```{r adt-background-diagnostic, eval=isTRUE(params$use_adt), fig.height=5, fig.width=8}
if (is.null(adt_matrix_raw)) {
  cat('No raw ADT matrix available; skipping ADT background diagnostic.\n')
} else {
  common_raw_barcodes <- intersect(colnames(raw_matrix), colnames(adt_matrix_raw))
  mt_rows <- rownames(raw_matrix)[grepl(mt_pattern, rownames(raw_matrix))]
  md <- data.frame(
    rna.size  = log10(pmax(Matrix::colSums(raw_matrix[, common_raw_barcodes, drop = FALSE]), 1)),
    prot.size = log10(pmax(Matrix::colSums(adt_matrix_raw[, common_raw_barcodes, drop = FALSE]), 1)),
    mt.prop   = Matrix::colSums(raw_matrix[mt_rows, common_raw_barcodes, drop = FALSE]) /
                pmax(Matrix::colSums(raw_matrix[, common_raw_barcodes, drop = FALSE]), 1),
    row.names = common_raw_barcodes
  )
  md$drop.class <- ifelse(rownames(md) %in% colnames(seurat), 'cell', 'background')

  ggplot(md, aes(x = prot.size, y = rna.size, color = mt.prop)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    theme_classic() +
    facet_wrap(~drop.class) +
    ggtitle('DSB background diagnostic: RNA vs ADT size by droplet class')
}
```

```{r adt-dsb-normalize, eval=isTRUE(params$use_adt)}
if (!requireNamespace('dsb', quietly = TRUE)) {
  cat('dsb package not installed; skipping ADT normalization.\n')
} else if (is.null(adt_matrix_raw)) {
  cat('No raw ADT matrix available; skipping ADT normalization.\n')
} else {
  common_raw_barcodes <- intersect(colnames(raw_matrix), colnames(adt_matrix_raw))
  rna.size  <- log10(pmax(Matrix::colSums(raw_matrix[, common_raw_barcodes, drop = FALSE]), 1))
  prot.size <- log10(pmax(Matrix::colSums(adt_matrix_raw[, common_raw_barcodes, drop = FALSE]), 1))
  is_background <- !(common_raw_barcodes %in% colnames(seurat))

  bg_rna_max  <- if (!is.na(params$dsb_background_rna_max))  params$dsb_background_rna_max  else stats::median(rna.size[is_background])
  bg_prot_min <- if (!is.na(params$dsb_background_prot_min)) params$dsb_background_prot_min else stats::quantile(prot.size[is_background], 0.01, names = FALSE)
  bg_prot_max <- if (!is.na(params$dsb_background_prot_max)) params$dsb_background_prot_max else stats::quantile(prot.size[is_background], 0.99, names = FALSE)

  background_barcodes <- common_raw_barcodes[
    is_background & rna.size <= bg_rna_max & prot.size >= bg_prot_min & prot.size <= bg_prot_max
  ]
  cat('DSB background droplets:', length(background_barcodes),
      '(rna.size <=', signif(bg_rna_max, 3), ', prot.size in [', signif(bg_prot_min, 3), ',', signif(bg_prot_max, 3), '])\n')

  if (length(background_barcodes) < 10) {
    cat('Fewer than 10 background droplets resolved; skipping DSB normalization.\n')
  } else {
    cell_protein_matrix <- as.matrix(GetAssayData(seurat, assay = 'ADT', layer = 'counts'))
    empty_drop_matrix   <- as.matrix(adt_matrix_raw[rownames(cell_protein_matrix), background_barcodes, drop = FALSE])
    isotype_controls    <- rownames(cell_protein_matrix)[grepl('Isotype', rownames(cell_protein_matrix), ignore.case = TRUE)]
    use_isotype_control <- length(isotype_controls) > 0

    dsb_out <- dsb::DSBNormalizeProtein(
      cell_protein_matrix     = cell_protein_matrix,
      empty_drop_matrix       = empty_drop_matrix,
      denoise.counts          = TRUE,
      use.isotype.control     = use_isotype_control,
      isotype.control.name.vec = if (use_isotype_control) isotype_controls else NULL
    )

    seurat[['ADT_DSB']] <- CreateAssayObject(data = dsb_out)
    cat('Isotype controls used for denoising:', if (use_isotype_control) paste(isotype_controls, collapse = ', ') else 'none detected', '\n')
    cat('ADT_DSB assay created with', nrow(dsb_out), 'features.\n')
  }
}
```

```{r adt-qc-plots, eval=isTRUE(params$use_adt) && ('ADT_DSB' %in% Assays(seurat)), fig.height=6, fig.width=10}
adt_features <- rownames(seurat[['ADT_DSB']])
raw_adt_totals <- Matrix::colSums(GetAssayData(seurat, assay = 'ADT', layer = 'counts'))

ggplot(data.frame(nCount_RNA = seurat$nCount_RNA, nCount_ADT = raw_adt_totals),
       aes(x = nCount_RNA, y = nCount_ADT)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_x_log10() + scale_y_log10() +
  theme_classic() +
  ggtitle('Total ADT UMI vs total RNA UMI per cell')

VlnPlot(seurat, features = adt_features[seq_len(min(6, length(adt_features)))],
        assay = 'ADT_DSB', layer = 'data', pt.size = 0, ncol = 3) &
  theme(legend.position = 'none')

DefaultAssay(seurat) <- 'ADT_DSB'
print(FeaturePlot(seurat, features = adt_features[seq_len(min(4, length(adt_features)))],
                   reduction = 'umap_rna', ncol = 2) & scale_color_viridis_c())
DefaultAssay(seurat) <- 'RNA'
```

```{r adt-cluster-heatmap, eval=isTRUE(params$use_adt) && ('ADT_DSB' %in% Assays(seurat)), fig.height=10, fig.width=8}
# Average DSB-normalized ADT expression per SCT cluster (no marker test needed);
# reuses the myColor/myBreaks palette defined in the marker-heatmap chunk above.
adt_avg <- AverageExpression(seurat, features = adt_features, assay = 'ADT_DSB', layer = 'data', group.by = 'seurat_clusters_sct')
adt_mat <- adt_avg[['ADT_DSB']]
if (nrow(adt_mat) > 0 && ncol(adt_mat) > 0) {
  pheatmap::pheatmap(
    adt_mat,
    scale = 'row',
    color = myColor,
    border_color = NULL,
    breaks = myBreaks,
    fontsize = 6,
    main = 'DSB-normalized average ADT expression per cluster (SCT clusters)'
  )
} else {
  cat('Empty ADT average expression matrix; skipping heatmap.\n')
}
```

# Doublet detection
Doublets are annotated but **not removed** here; reassess after integration/merging. Doublets per RNA (log-normalized) clusters are shown below.

```{r doublet-detection}
# Run scDblFinder twice (with RNA cluster labels and with clusters = NULL), then
# save both annotations in metadata. Doublets are flagged but NOT removed here;
# reassess after integration/merging.
seurat_sce_clustered <- as.SingleCellExperiment(seurat)
seurat_sce_clustered$cluster_for_dbl <- as.character(seurat$seurat_clusters_rna)
seurat_sce_clustered <- scDblFinder(seurat_sce_clustered, clusters = 'cluster_for_dbl')

seurat_sce_unclustered <- as.SingleCellExperiment(seurat)
seurat_sce_unclustered <- scDblFinder(seurat_sce_unclustered, clusters = NULL)

seurat$doublet_class_clustered <- seurat_sce_clustered$scDblFinder.class
seurat$doublet_score_clustered <- seurat_sce_clustered$scDblFinder.score
seurat$doublet_class_unclustered <- seurat_sce_unclustered$scDblFinder.class
seurat$doublet_score_unclustered <- seurat_sce_unclustered$scDblFinder.score

# Keep compatibility columns mapped to the clustered run.
seurat$doublet_class <- seurat$doublet_class_clustered
seurat$doublet_score <- seurat$doublet_score_clustered

rds_path <- file.path(params$output_dir, paste0(params$sample_name, '_QC.rds'))
saveRDS(seurat, file = rds_path)
cat('Saved:', rds_path, '\n')
```

```{r doublet-plots}
# Doublet distribution across RNA clusters for both scDblFinder runs
cat('Clustered run (clusters = seurat_clusters_rna):\n')
print(table(seurat$seurat_clusters_rna, seurat$doublet_class_clustered))
cat('\nUnclustered run (clusters = NULL):\n')
print(table(seurat$seurat_clusters_rna, seurat$doublet_class_unclustered))

Idents(seurat) <- 'doublet_class_clustered'
DimPlot(seurat, reduction = 'umap_rna') +
  scale_color_manual(values = c('singlet' = 'grey70', 'doublet' = 'red3')) +
  ggtitle('scDblFinder (clustered): doublet class')
FeaturePlot(seurat, features = 'doublet_score_clustered', reduction = 'umap_rna') +
  ggtitle('scDblFinder (clustered): doublet score')

Idents(seurat) <- 'doublet_class_unclustered'
DimPlot(seurat, reduction = 'umap_rna') +
  scale_color_manual(values = c('singlet' = 'grey70', 'doublet' = 'red3')) +
  ggtitle('scDblFinder (unclustered): doublet class')
FeaturePlot(seurat, features = 'doublet_score_unclustered', reduction = 'umap_rna') +
  ggtitle('scDblFinder (unclustered): doublet score')

Idents(seurat) <- 'seurat_clusters_rna'
VlnPlot(
  seurat,
  features = c('doublet_score_clustered', 'doublet_score_unclustered')
) + NoLegend()

# Compare UMI count distributions between singlets and doublets for each run.
doublet_umi_df <- seurat@meta.data[, c(
  'nCount_RNA',
  'doublet_class_clustered',
  'doublet_class_unclustered'
)]

ggplot(doublet_umi_df, aes(x = doublet_class_clustered, y = nCount_RNA, fill = doublet_class_clustered)) +
  geom_violin(trim = FALSE) +
  scale_y_log10() +
  scale_fill_manual(values = c('singlet' = 'grey70', 'doublet' = 'red3')) +
  theme_classic() +
  theme(legend.position = 'none') +
  xlab('scDblFinder class') +
  ylab('UMI count (nCount_RNA, log10 scale)') +
  ggtitle('UMI count by doublet class (clustered scDblFinder)')

ggplot(doublet_umi_df, aes(x = doublet_class_unclustered, y = nCount_RNA, fill = doublet_class_unclustered)) +
  geom_violin(trim = FALSE) +
  scale_y_log10() +
  scale_fill_manual(values = c('singlet' = 'grey70', 'doublet' = 'red3')) +
  theme_classic() +
  theme(legend.position = 'none') +
  xlab('scDblFinder class') +
  ylab('UMI count (nCount_RNA, log10 scale)') +
  ggtitle('UMI count by doublet class (unclustered scDblFinder)')

# Proportion of doublets per cluster for each run
prop.doublets.clustered <- as.data.frame(
  prop.table(table(seurat$seurat_clusters_rna, seurat$doublet_class_clustered), margin = 1)
)
ggplot(prop.doublets.clustered, aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = 'identity') +
  theme_classic() +
  xlab('Cluster number') +
  ylab('Fraction doublet class') +
  ggtitle('Doublet fraction by cluster (clustered scDblFinder)')

prop.doublets.unclustered <- as.data.frame(
  prop.table(table(seurat$seurat_clusters_rna, seurat$doublet_class_unclustered), margin = 1)
)
ggplot(prop.doublets.unclustered, aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = 'identity') +
  theme_classic() +
  xlab('Cluster number') +
  ylab('Fraction doublet class') +
  ggtitle('Doublet fraction by cluster (unclustered scDblFinder)')
```

